---
output: github_document
bibliography: inst/REFERENCES.bib
---
  
  ```{r, echo = FALSE, message=F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = "png",
  dpi = 200,
  fig.align = "center",
  knitr::opts_chunk$set(comment = NA)
)
library(bgms)
```

# bgms: Bayesian Analysis of Graphical Models
  
<p style="text-align: justify">The `R` package **bgms** provides tools for a 
Bayesian analysis of graphical models describing networks of variables. The 
package uses Bayesian variable selection methods to model the underlying network 
structure. The methods are organized around two general approaches for Bayesian 
variable selection: (1) EM variable selection and (2) Gibbs variable selection. 
The key distinction is that the former uses a continuous spike and slab prior 
distribution on the pairwise interactions [@MarsmanEtAl_2022_objective] that 
allows us to use EM variable selection [@RockovaGeorge_2014]. The Gibbs variable 
selection approach [@GeorgeMcCulloch_1993], on the other hand, stipulates a 
discrete spike and slab prior on the pairwise interactions, which allows us to 
set the interactions to exact zeroes. To account for the discontinuity at zero, 
we embed a Metropolis approach for mixtures of mutually singular distributions 
[@GottardoRaftery_2008] in a Gibbs sampler.</p>


## Why use Markov Random Fields?
<p style="text-align: justify">Multivariate analysis using graphical models has 
received much attention in the recent psychological and psychometric literature 
[@RobinaughEtAl_2020, @MarsmanRhemtulla_2022_SIintro, @Steinley_2021_SIintro, 
@ContrerasEtAl_2019]. Most of these graphical models are Markov Random Field 
(MRF) models, whose graph structure reflects the conditional associations 
between variables [@KindermannSnell1980]. In these models, a missing edge 
between two variables in the network implies that these variables are 
independent, given the remaining variables [@Lauritzen2004]. In other words, the 
remaining variables of the network fully account for the potential association 
between the unconnected variables.</p>

## Why use a Bayesian approach to analyze the MRF?
<p style="text-align: justify">Testing the structure of the MRF requires us to 
determine the plausibility of the opposing hypotheses of conditional dependence 
and conditional independence. For example, how plausible are network structures 
that include the edge between variables 3 and 9 compared to network structures 
that exclude this edge? Frequentist approaches are limited in this respect, 
because they can only reject the conditional independence hypothesis, but not 
support it [@WagenmakersEtAl_2018_BIP1, @Wagenmakers_2007]. This creates the 
problem that, if an edge is excluded, we do not know whether this is because the 
edge is absent in the population, or because we lack the power to reject the 
null hypothesis of independence.  To avoid this problem, we will use a Bayesian 
approach using Bayes factors [@KassRaftery_1995]. The inclusion Bayes factor 
[@HuthEtAl_2023] allows us to quantify how much the data support both 
conditional dependence ---<em>evidence of edge presence</em>--- or conditional 
independence ---<em>evidence of edge absence</em>. It also allows us to conclude 
that there is only limited support for either hypothesis [@Dienes_2014] ---an 
<em>absence of evidence</em>.</p>


## Installation

The current developmental version can be installed with	
```{r gh-installation, eval = FALSE}	
if (!requireNamespace("remotes")) {	
  install.packages("remotes")	
}	
remotes::install_github("MaartenMarsman/bgms")
```


## References