---
title: "Checking your fitted model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking your fitted model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.4
)
```

# Introduction

A fitted network raises questions of two kinds. *Which edges does the evidence
settle?* is a question about the posterior, and `verdicts()` answers it. *Does
the model describe the data?* is a question about fit, and it splits in two:
whether the conditional prediction for each variable is calibrated, which
`calibration_check()` answers, and whether the joint distribution is
reproduced, which you check by building a display on `simulate()`.

This vignette works through all of them on the Wenchuan data. The convergence
diagnostics -- R-hat, effective sample size, the trust gauge -- are a separate
matter and live in the diagnostics vignette.

```{r, eval=FALSE}
library(bgms)
data = Wenchuan[, 1:9]
fit = bgm(data, seed = 1234)
```

```{r, include=FALSE}
library(bgms)
data = Wenchuan[, 1:9]
fit = bgm(data,
  seed = 1234, chains = 2, iter = 1000, warmup = 1000,
  display_progress = "none", verbose = FALSE
)
```

# What the evidence settles

Each edge carries an inclusion Bayes factor, and reading it at a threshold
gives one of three verdicts: evidence of presence, evidence of absence, or
undecided. The third is not a failure of the analysis; it is the honest answer
when the data do not separate the two hypotheses.

```{r}
edges = verdicts(fit)
edges
```

The `fragile` column is the part that cannot be read off the Bayes factor
alone. A verdict is fragile when a threshold sits within two standard errors of
the estimated evidence, which is exactly the regime in which Monte Carlo noise,
not the data, decides the answer. In a known-truth calibration study of 37,010
edge-fits across ordinal, binary, and Gaussian graphical models, every verdict
error sat within 0.25 of a threshold on the log10 Bayes factor scale, and no
edge further out was ever misclassified.

A fragile verdict is not a wrong verdict; it is a verdict the run is too short
to settle. The remedy is more iterations, not a different model.

```{r}
edges[edges$fragile, c("parameter", "log10_bf", "verdict")]
```

The default picture encodes the same three verdicts, so what a reader sees and
what the table says cannot drift apart.

```{r, fig.height = 5.5, fig.width = 6}
plot(fit)
```

Solid edges have evidence of presence, with width scaled by the model-averaged
weight; dotted grey edges are undecided; edges with evidence of absence are not
drawn at all. For one edge in detail, the spike-and-slab posterior shows how
the mass is actually divided:

```{r, fig.height = 3.6, fig.width = 5}
plot_edge_posterior(fit, "intrusion", "upset")
```

Centrality is a function of the network, so it inherits the same uncertainty.
Evaluating it on every posterior draw gives its posterior directly, which is
what `extract_centrality()` returns.

```{r, fig.height = 3.6, fig.width = 6}
strength = extract_centrality(fit)
head(summary(strength))
plot(strength)
```

# Are the conditional predictions calibrated?

The model's person-level predictions are the conditional distributions
`P(x_ij | the rest of case i's responses)`, which is the model's own regression
unit. `calibration_check()` compares those predictions with what was observed, one
variable at a time: an isotonic fit of observed frequency on predicted
probability for a discrete variable, and the probability integral transform
for a continuous one.

```{r}
calibration = calibration_check(fit, nrep = 200, seed = 1)
calibration
```

One panel per variable is not a display preference. Pooling across variables
lets a variable predicted too high and one predicted too low cancel, and the
pooled curve then tracks the diagonal while neither variable does.

```{r, fig.height = 5.5, fig.width = 7}
plot(calibration)
```

The grey band is the wander of a model that is calibrated by construction. It
is built by resampling each case's *category* from its own predicted
distribution, rather than resampling the threshold events independently: the
cumulative events of one case are nested, and treating them as independent
would make the band too narrow and an ordinary wander look like
miscalibration.

A continuous variable has no category to have fallen in, so it gets the same
question through the probability integral transform: `u = F(y | rest)`, which is
uniform exactly when the conditional predictive distribution is right. Its
panel is the distribution of those `u` values against the uniform diagonal,
and `F` is the mixture over posterior draws, so the parameter uncertainty is
inside the distribution the observation is transformed by. That band is not
resampled from the model at all -- under the transform the null is uniform
whatever the density was -- so it depends only on the number of cases. A mixed
fit draws both panel kinds on one figure and records which is which in the
`kind` column.

# Checking the joint distribution with simulate()

Calibrated conditional predictions do not imply that the model reproduces the
joint distribution: predicting each variable from its neighbours is an easier
task than reproducing how all of them covary, and a model can do the first
while falling short on the second.

bgms ships no joint-level check, and the omission is deliberate. The natural
candidates -- category proportions, pairwise correlations, sum scores -- are
close to quantities the model already fits, so predictive intervals around
them usually cover, and a check that usually passes is reassurance rather than
evidence. What does have diagnostic value is reading a display for systematic
pattern, and that is an analyst's judgement rather than a pass/fail statistic.

`simulate()` gives you the material. One replicated dataset per posterior draw
propagates both parameter and structure uncertainty into the replicates:

```{r}
observed = as.matrix(data[complete.cases(data), ])

replicates = simulate(fit,
  nsim = nrow(observed), method = "posterior-sample",
  ndraws = 200, seed = 1, display_progress = "none"
)

lower_triangle = lower.tri(matrix(0, ncol(observed), ncol(observed)))
observed_rho = cor(observed, method = "spearman")[lower_triangle]
replicated_rho = vapply(
  replicates,
  function(x) cor(x, method = "spearman")[lower_triangle],
  observed_rho
)
```

Plot the observed value against the predictive distribution rather than
counting how often the two agree. The count is the weak reading; the
informative one is whether the points depart from the identity line in a
direction:

```{r, fig.height = 3.6, fig.width = 4.6}
predicted_rho = rowMeans(replicated_rho)
plot(predicted_rho, observed_rho,
  xlab = "Posterior predictive Spearman correlation",
  ylab = "Observed Spearman correlation",
  pch = 16, col = "grey25", cex = 0.8, las = 1,
  xlim = range(c(predicted_rho, observed_rho)),
  ylim = range(c(predicted_rho, observed_rho))
)
abline(0, 1, col = "grey55")

mean(observed_rho > apply(replicated_rho, 1, median))
```

Every one of the observed correlations sits above its predictive median, and
the cloud tilts away from the identity line at the low end: the model
understates the weaker associations. Under a model that reproduced the
dependence, each correlation would fall above or below its median at about
even odds, so a clean sweep is not something Monte Carlo error produces. That
is a real finding, and no interval-coverage summary would have delivered it --
the tilt is visible because the display shows observed against predicted, one
point per pair.

Any statistic works the same way: replace the correlation with a sum-score
distribution, a tail probability, or whatever the substantive question turns
on.

# What the three checks answer

They are three different questions, and answering one says nothing about the
other two.

- `verdicts()`: which edges the data settle, and which verdicts the run is too
  short to settle.
- `calibration_check()`: whether to trust a prediction for one variable.
- A `simulate()`-based display: whether to trust the network as a description
  of the joint distribution.

An edge can be firmly settled in a model that reproduces the joint
distribution poorly, and a model can predict each variable well while
understating the dependence among them.

# Next steps

- The diagnostics vignette covers convergence: R-hat, effective sample size,
  the NUTS diagnostics, and the hierarchical prior trust gauge.
- `prior_sensitivity_check()` reports whether a verdict depends on the prior
  rather than on the run length, which is the other way a verdict can be
  unstable.
