---
title: "Checking your fitted model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking your fitted model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.4
)
```

# Introduction

A fitted network answers two different questions, and they need two different
checks. *Which edges does the evidence settle?* is a question about the
posterior, and `verdicts()` answers it. *Does the model reproduce the data?*
is a question about fit, and the two predictive checks answer it.

This vignette runs all three on the Wenchuan data. The convergence
diagnostics -- R-hat, effective sample size, the trust gauge -- are a separate
matter and live in the diagnostics vignette.

```{r, eval=FALSE}
library(bgms)
data = Wenchuan[, 1:9]
fit = bgm(data, seed = 1234)
```

```{r, include=FALSE}
library(bgms)
data = Wenchuan[, 1:9]
fit = bgm(data,
  seed = 1234, chains = 2, iter = 1000, warmup = 1000,
  display_progress = "none", verbose = FALSE
)
```

# What the evidence settles

Each edge carries an inclusion Bayes factor, and reading it at a threshold
gives one of three verdicts: evidence of presence, evidence of absence, or
undecided. The third is not a failure of the analysis; it is the honest answer
when the data do not separate the two hypotheses.

```{r}
edges = verdicts(fit)
edges
```

The `fragile` column is the part that cannot be read off the Bayes factor
alone. A verdict is fragile when a threshold sits within two standard errors of
the estimated evidence, which is exactly the regime in which Monte Carlo noise,
not the data, decides the answer. In a known-truth calibration study of 37,010
edge-fits across ordinal, binary, and Gaussian graphical models, every verdict
error sat within 0.25 of a threshold on the log10 Bayes factor scale, and no
edge further out was ever misclassified.

A fragile verdict is not a wrong verdict; it is a verdict the run is too short
to settle. The remedy is more iterations, not a different model.

```{r}
edges[edges$fragile, c("parameter", "log10_bf", "verdict")]
```

The default picture encodes the same three verdicts, so what a reader sees and
what the table says cannot drift apart.

```{r, fig.height = 5.5, fig.width = 6}
plot(fit)
```

Solid edges have evidence of presence, with width scaled by the model-averaged
weight; dotted grey edges are undecided; edges with evidence of absence are not
drawn at all. For one edge in detail, the spike-and-slab posterior shows how
the mass is actually divided:

```{r, fig.height = 3.6, fig.width = 5}
plot_edge_posterior(fit, "intrusion", "upset")
```

Centrality is a function of the network, so it inherits the same uncertainty.
Evaluating it on every posterior draw gives its posterior directly, which is
what `extract_centrality()` returns.

```{r, fig.height = 3.6, fig.width = 6}
strength = extract_centrality(fit)
head(summary(strength))
plot(strength)
```

# Does the model reproduce the data?

`posterior_predictive_check()` draws one replicated dataset per posterior
draw, so uncertainty about both the parameters and the structure propagates
into the replicates, and compares three statistics with their predictive
intervals.

```{r}
ppc = posterior_predictive_check(fit, ndraws = 200, seed = 1, display_progress = "none")
ppc
```

The statistics run from local to global. The category proportions involve one
variable at a time, the Spearman correlations two, and the sum-score
distribution depends on how all nine variables depend on one another. A model
can pass the first and fail the last.

```{r, fig.height = 3.2, fig.width = 8}
plot(ppc)
```

Read this on two layers.

The first is **coverage**: how many elements fall outside their interval. With
95% intervals about one element in twenty lands outside even when the model
fits, which is why each panel's subtitle states the number a calibrated model
is expected to cover. A handful of accent-coloured points is the expected
picture.

The second is **pattern**, and it is the layer a coverage count cannot show. In
the pairwise panel the observed correlations sit above the identity line
almost without exception at the low end: the model systematically understates
the weaker associations. That is misfit even where the intervals happen to
cover, and it is visible only because the panel plots observed against
predicted rather than reporting a single number.

# Are the conditional predictions calibrated?

The model's person-level predictions are the conditional distributions
`P(x_ij | the rest of case i's responses)`, which is the model's own regression
unit. `calibration_check()` compares those predicted probabilities with the
observed frequencies through an isotonic fit, one variable at a time.

```{r}
calibration = calibration_check(fit, nrep = 200, seed = 1)
calibration
```

One panel per variable is not a display preference. Pooling across variables
lets a variable predicted too high and one predicted too low cancel, and the
pooled curve then tracks the diagonal while neither variable does.

```{r, fig.height = 5.5, fig.width = 7}
plot(calibration)
```

The grey band is the wander of a model that is calibrated by construction. It
is built by resampling each case's *category* from its own predicted
distribution, rather than resampling the threshold events independently: the
cumulative events of one case are nested, and treating them as independent
would make the band too narrow and an ordinary wander look like
miscalibration.

# Reading the two checks together

The two checks find different things on these data, which is the reason to run
both.

The calibration curves stay near the diagonal and wander in and out of their
bands, with the departures concentrated in a few variables (`avoidth` and
`avoidact` here) and mild in the rest. Read alongside the per-variable panels,
that is a model whose conditional predictions are usable but not uniformly
well calibrated.

The predictive check is more clear-cut, and it points somewhere else: the
pairwise panel covers well below expectation and tilts systematically, so the
model understates how strongly the variables covary. Predicting each variable
from its neighbours is an easier task than reproducing how all of them covary,
and a model can do reasonably at the first while falling short on the second.

So the checks answer different questions. The calibration check bears on
whether to trust a prediction for one variable; the predictive check bears on
whether to trust the network as a description of the joint distribution. And
`verdicts()` bears on which of its edges the data actually settle, which is a
third question again -- an edge can be firmly settled in a model that
reproduces the joint distribution poorly.

# Next steps

- The diagnostics vignette covers convergence: R-hat, effective sample size,
  the NUTS diagnostics, and the hierarchical prior trust gauge.
- `prior_sensitivity_check()` reports whether a verdict depends on the prior
  rather than on the run length, which is the other way a verdict can be
  unstable.
