---
title: "Diagnostics and Spike-and-Slab Summaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics and Spike-and-Slab Summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Introduction

This vignette illustrates how to inspect convergence diagnostics and how
to interpret spike-and-slab summaries in **bgms** models. For some of the model
variables spike-and-slab priors introduce binary indicator variables 
that govern whether the effect is included or not. Their posterior distributions 
can be summarized with inclusion probabilities and Bayes factors.

# Example fit

We use a subset of the Wenchuan dataset:

```{r, eval=FALSE}
library(bgms)
data = Wenchuan[, 1:5]
fit = bgm(data, seed = 1234)
```

```{r, include=FALSE}
library(bgms)
data = Wenchuan[, 1:5]
fit = bgm(data, seed = 1234, chains = 2, display_progress = "none", verbose = FALSE)
```

# Convergence diagnostics

The quality of the Markov chain can be assessed with common MCMC diagnostics:

```{r}
summary(fit)$pairwise
```

- R-hat values close to 1 (typically below 1.01) suggest convergence [@VehtariEtAl_2021].
- The effective sample size (ESS) reflects the number of independent samples that would provide equivalent precision. Larger ESS values indicate more reliable estimates.
- The Monte Carlo standard error (MCSE) measures the additional variability introduced by using a finite number of MCMC draws. A small MCSE relative to the posterior standard deviation indicates stable estimates, whereas a large MCSE suggests that more samples are needed.

## Two ESS measures for edge-selected parameters

With edge or difference selection active, the effect parameters are governed by
spike-and-slab priors. The corresponding parameter is set to exactly zero when 
the effect is excluded, rather than being removed from the model. Because the 
parameter has a well-defined value at every iteration, the full chain — 
including zeros — is a valid sequence for computing ESS.

- **n_eff** is the unconditional ESS, computed from the full effect chain. It
  measures how precisely the overall posterior mean is estimated.
- **n_eff_mixt** is the mixture ESS. It measures how precisely the posterior
  mean of the effect is estimated while accounting for the additional
  uncertainty introduced by the spike-and-slab selection. When the indicator
  rarely switches between inclusion and exclusion (fewer than 5 transitions),
  `n_eff_mixt` is suppressed in the printed output.

## Traceplots

Users can inspect traceplots by extracting raw samples directly. Here is an 
example for the pairwise effect parameter.

```{r, fig.width= 7, fig.height= 7}
param_index = 1
chains = fit$raw_samples$pairwise
nchains = length(chains)
cols = c("firebrick", "steelblue", "darkgreen", "goldenrod")

plot(chains[[1]][, param_index],
  type = "l", col = cols[1],
  xlab = "Iteration", ylab = "Value",
  main = "Traceplot of pairwise[1]",
  ylim = range(sapply(chains, function(ch) range(ch[, param_index])))
)
if(nchains > 1) {
  for(c in 2:nchains) {
    lines(chains[[c]][, param_index], col = cols[c])
  }
}
```


# Spike-and-slab summaries

The spike-and-slab prior yields posterior inclusion probabilities for
edges:

```{r}
coef(fit)$indicator
```
- Values near 1.0: strong evidence the edge is present.
- Values near 0.0: strong evidence the edge is absent.
- Values near 0.5: inconclusive (absence of evidence).

# Bayes factors

When the prior inclusion probability for an edge is equal to 0.5 (e.g., using
`edge_prior = bernoulli_prior(0.5)`, the default, or a symmetric Beta prior
`edge_prior = beta_bernoulli_prior(alpha, beta)` with `alpha == beta`), we can
directly transform inclusion probabilities into
Bayes factors for edge presence vs absence:

```{r}
# Example for one edge
p = coef(fit)$indicator[1, 5]
BF_10 = p / (1 - p)
BF_10
```

Here the Bayes factor in favor of inclusion (H1) is small, meaning that there is little evidence for inclusion. Since the Bayes factor is transitive, we can 
use it to express the evidence in favor of exclusion (H0) as

```{r}
1 / BF_10
```
This Bayes factor shows that there is strong evidence for the absence of a 
network relation between the variables `intrusion` and `physior`.

# NUTS diagnostics

When using `update_method = "nuts"` (the default), additional diagnostics are 
available to assess the quality of the Hamiltonian Monte Carlo sampling. These
can be accessed via `fit$nuts_diag`:

```{r}
fit$nuts_diag$summary
```

## E-BFMI

E-BFMI (Energy Bayesian Fraction of Missing Information) measures how  efficiently the sampler explores the posterior. It compares the typical size of energy changes between successive samples to the overall spread of energies. Values close to 1 indicate that the sampler moves freely across the energy landscape; values below 0.3 suggest the sampler may be getting stuck or that the chain has not yet settled into its stationary distribution.

A low E-BFMI does not necessarily mean your results are wrong, but it does warrant further investigation. In models with edge selection, the most common cause is that the warmup period was too short for the discrete graph structure to equilibrate. Increasing `warmup` often resolves this.

## Divergent transitions

Divergent transitions occur when the numerical integrator encounters regions of the posterior where the curvature changes too rapidly for the current step size. A small number of divergences (say, fewer than 0.1% of samples) is generally acceptable. However, many divergences indicate that the sampler may be missing important parts of the posterior.

If you see a large number of divergences, consider increasing `target_accept` (which makes the sampler use a smaller step size) and, if this does not fix it, switching to `update_method = "adaptive-metropolis"`.

## Tree depth

NUTS builds trajectories by repeatedly doubling their length until a "U-turn" criterion is satisfied. If the trajectory frequently reaches the maximum allowed depth (`nuts_max_depth`, default 10), it suggests the sampler may benefit from longer trajectories to explore the posterior efficiently. Hitting the maximum depth occasionally is normal; hitting it on most iterations may indicate challenging posterior geometry. If this happens, consider increasing `nuts_max_depth`.

## Non-reversible steps

For MRFs with continuous variables, the leapfrog integrator enforces equality constraints through a projection step. After each forward step, the integrator checks whether reversing the step returns to the starting point. When the round-trip error exceeds a tolerance scaled by the square of the step size, the step is flagged as non-reversible.

A small number of non-reversible steps is not a concern. A large number indicates that the step size is too large for the constraint geometry. Because the step size is tuned during warmup, the most effective remedy is to increase `warmup` so the adapter has more time to find an appropriate step size. If non-reversible steps persist after increasing warmup, switch to `update_method = "adaptive-metropolis"`.

## Warmup and equilibration

Standard HMC/NUTS warmup is designed to tune the step size and mass matrix for the continuous parameters. In models with edge selection, the discrete graph structure may take longer to reach its stationary distribution than the continuous parameters. As a result, even after warmup completes, the first portion of the sampling phase may still show transient behavior (i.e., non-stationarity).
  
The `warmup_check` component provides simple diagnostics that compare the first and second halves of the post-warmup samples:

```{r}
fit$nuts_diag$warmup_check
```

The returned list contains the following fields (one value per chain):

- **warmup_incomplete**: A logical flag that is `TRUE` when any of the indicators below suggest the chain may not have reached stationarity.
- **energy_slope**: The slope of a linear regression of energy against iteration number. A slope near zero indicates stable energy; a significant negative slope suggests the chain is still drifting toward higher-probability regions.
- **slope_significant**: `TRUE` if the energy slope is statistically significant (p < 0.01).
- **ebfmi_first_half** and **ebfmi_second_half**: E-BFMI computed separately for the first and second halves of the post-warmup samples. If the first-half value is much lower (for example, below 0.3) while the second-half value is healthy, the early samples were likely still settling.
- **var_ratio**: The ratio of energy variance in the first half to that in the second half. A ratio much greater than 1 (for example, above 2) indicates higher variability early on, consistent with transient behavior.

If these diagnostics suggest the chain was still settling, increase `warmup` and re-run the model. If diagnostics remain problematic after a substantial increase (for example, doubling or tripling `warmup`), consider re-fitting with `update_method = "adaptive-metropolis"` and comparing the posterior summaries. If the two samplers produce similar results, the estimates are likely trustworthy despite the warnings; if they differ substantially, that warrants further investigation of the model or data.

# Hierarchical prior trust gauge

When you use the hierarchical graph prior (`precision_graph_prior = "hierarchical"`), the sampler must weigh, for every candidate edge, how the model's normalizing constant changes when that edge is added or removed. Computing this exactly at every step would be prohibitively slow, so the sampler uses a fast approximation in its place. The approximation is accurate across the prior scales and model sizes it was calibrated on, but it can drift outside that range, most often when there are many variables or a very wide prior scale. When it drifts far enough, it starts to influence which edges the sampler includes, so the recovered network reflects the shortcut rather than the data and prior alone. The trust gauge exists to tell you whether that is happening in your fit.

## The two alarms

While the sampler runs, the gauge audits its own decisions. On a subset of the edge moves the chain actually makes, it redoes the add/remove decision the exact (slow) way and checks whether the outcome would have been different. Auditing the chain's own moves is deliberate: it measures the approximation where the sampler is actually spending its time, rather than at arbitrary test configurations. The audit is only produced under the hierarchical prior, and its summary is stored in `fit$zratio_diag`, with one row per chain.

The gauge reports two complementary alarms. The first is `flip_rate`, the fraction of audited decisions that would have come out differently under the exact calculation. A chain is flagged on this channel above a 1% tolerance, after allowing for the small amount of Monte Carlo noise in the exact reference itself. Because the flip rate counts decisions rather than measuring a distance in abstract units, the 1% tolerance means the same thing regardless of the number of variables or the choice of prior.

The flip rate only registers error that lands near an accept/reject boundary, so a small but consistent error can pass it silently while still shifting the recovered inclusion probabilities: applied to every move a chain makes, a consistent lean relocates where the chain settles even when no single decision visibly changes. The second channel, `harm_pred`, projects the measured error onto that scale directly: it averages each audited edge's signed error weighted by that edge's own inclusion sensitivity (an error on an edge the data have pinned near 0 or 1 cannot move the result and does not count), then multiplies by the feedback amplification of the edge prior (learned inclusion probabilities amplify a consistent lean; a fixed inclusion probability does not). The result is the first-order predicted distortion of the mean posterior inclusion probability, flagged when it exceeds 0.01 and the weighted error is resolved above its own Monte Carlo uncertainty. It is a targeted alarm for a coherent lean, not a bound on all approximation error: edge-specific distortions that cancel in the mean are outside its scope. This channel is computed for the Bernoulli and Beta-Bernoulli edge priors; under other priors it is reported as `NA`.

The `per_chain` data frame in `fit$zratio_diag` records the following fields (one row per chain):

- **flip_rate**: The fraction of audited decisions that would have come out differently under the exact calculation. Together with `harm_pred`, one of the two fields to watch.
- **flag**: `TRUE` when the chain's flip rate exceeds the 1% tolerance after allowing for reference noise.
- **noise_floor**: How much of the flip rate could be explained by Monte Carlo noise in the exact reference alone. The flag compares `flip_rate` against the tolerance plus this floor, so a chain is only flagged when the effect is larger than the reference's own uncertainty.
- **se_mean**: The average difference between the approximate and exact edge calculations, measured on the log scale the sampler works on. A value near zero means the approximation is unbiased on this chain's moves; a value consistently away from zero means it leans in one direction. An approximation can lean this way without changing any decisions when the data strongly favor one outcome; the harm channel exists to translate exactly that situation into consequences for the recovered network.
- **se_sd**: The spread of that difference across the audited moves.
- **se_mcse**, **se_se**: The Monte Carlo uncertainty of `se_mean`: `se_mcse` is the contribution of the exact reference's own noise, and `se_se` combines it with the spread across audited moves. The harm channel only flags a lean that is resolved above twice its own uncertainty (accounting for repeated audits of the same edge and for the reference noise), so a flag reflects a measured effect rather than noise.
- **amplification**, **kappa**, **harm_pred**, **harm_flag**: The harm channel. `amplification` is the feedback factor estimated from the chain's own inclusion probabilities and the edge prior; `kappa` is the predicted shift of the mean inclusion probability per nat of coherent error (a regime summary: how exposed this fit is to approximation error, independent of how large the error turned out to be); `harm_pred` is the projected distortion itself; `harm_flag` is `TRUE` when that projection exceeds 0.01 and is resolved. A fit can be flagged here with a quiet flip rate: that combination means the error is not changing individual decisions but is consistently leaning the recovered network in one direction.
- **n_ent**, **n_ref**, **n_capped**: How much the audit rests on. `n_ent` is the number of non-trivial edge moves the gauge saw, `n_ref` the number it actually checked against the exact calculation, and `n_capped` the number left unchecked because a per-sweep budget was reached. A small `n_ref` means both channels are based on few comparisons and should be read as a rough indication.

## What to do when a chain is flagged

A flagged chain tells you the approximation is affecting that chain's recovered structure — a `flag` on the flip channel means individual edge decisions are changing; a `harm_flag` means a consistent lean is shifting the inclusion probabilities — so the structure should be treated with caution. The approximation is calibrated online during an appended warm-up window before sampling begins, and the length of that window is controlled by `calibration_window`. Left at its default (`NULL`), the window is empty for fewer than fifteen variables and runs for 15% of `warmup` at fifteen or more, which is why flags appear mainly in larger models. The most effective remedy is to lengthen that window by setting `calibration_window` to a larger value, which gives the calibration more sweeps to settle; this is the first thing to try.

If a flag persists after a substantial increase in `calibration_window`, narrow the prior scale toward its calibrated range, since a very wide scale is the other common cause of drift. Failing that, you can keep the recovered structure but treat the flagged edges as provisional, confirming them another way, for example by re-fitting the neighborhood of those edges with a narrower prior and checking that they survive.

For evidence-free fits (prior-only sampling, or data too weak to pin the edges) under a Beta-Bernoulli edge prior, a `harm_pred` flag deserves extra weight: the learned inclusion probability amplifies a small per-edge lean, and lengthening the calibration window may not remove the residual. In that setting prefer the joint prior specification, which computes the edge weights without this approximation. A direct check is also available there: under either specification the average posterior inclusion probability of a prior-only fit must equal the edge prior's mean, which for `beta_bernoulli_prior(a, b)` is `a / (a + b)`.

# Next steps

- See *Getting Started* for a simple one-sample workflow.
- See *Model Comparison* for group differences.
