---
title: "Diagnostics and Spike-and-Slab Summaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics and Spike-and-Slab Summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Introduction

This vignette illustrates how to inspect convergence diagnostics and how
to interpret spike-and-slab summaries in **bgms** models. For some of the model
variables spike-and-slab priors introduce binary indicator variables 
that govern whether the effect is included or not. Their posterior distributions 
can be summarized with inclusion probabilities and Bayes factors.

# Example fit

We use a subset of the Wenchuan dataset:

```{r, eval=FALSE}
library(bgms)
data = Wenchuan[, 1:5]
fit = bgm(data, seed = 1234)
```

```{r, include=FALSE}
library(bgms)
data = Wenchuan[, 1:5]
fit = bgm(data, seed = 1234, chains = 2, display_progress = "none", verbose = FALSE)
```

# Convergence diagnostics

The quality of the Markov chain can be assessed with common MCMC diagnostics:

```{r}
summary(fit)$pairwise
```

- R-hat values close to 1 (typically below 1.01) suggest convergence [@VehtariEtAl_2021].
- The effective sample size (ESS) reflects the number of independent samples that would provide equivalent precision. Larger ESS values indicate more reliable estimates.
- The Monte Carlo standard error (MCSE) measures the additional variability introduced by using a finite number of MCMC draws. A small MCSE relative to the posterior standard deviation indicates stable estimates, whereas a large MCSE suggests that more samples are needed.

## Two ESS measures for edge-selected parameters

With edge or difference selection active, the effect parameters are governed by
spike-and-slab priors. The corresponding parameter is set to exactly zero when 
the effect is excluded, rather than being removed from the model. Because the 
parameter has a well-defined value at every iteration, the full chain — 
including zeros — is a valid sequence for computing ESS.

- **n_eff** is the unconditional ESS, computed from the full effect chain. It
  measures how precisely the overall posterior mean is estimated.
- **n_eff_mixt** is the mixture ESS. It measures how precisely the posterior
  mean of the effect is estimated while accounting for the additional
  uncertainty introduced by the spike-and-slab selection. When the indicator
  rarely switches between inclusion and exclusion (fewer than 5 transitions),
  `n_eff_mixt` is suppressed in the printed output.

## Traceplots

Users can inspect traceplots by extracting raw samples directly. Here is an 
example for the pairwise effect parameter.

```{r, fig.width= 7, fig.height= 7}
param_index = 1
chains = fit$raw_samples$pairwise
nchains = length(chains)
cols = c("firebrick", "steelblue", "darkgreen", "goldenrod")

plot(chains[[1]][, param_index],
  type = "l", col = cols[1],
  xlab = "Iteration", ylab = "Value",
  main = "Traceplot of pairwise[1]",
  ylim = range(sapply(chains, function(ch) range(ch[, param_index])))
)
if(nchains > 1) {
  for(c in 2:nchains) {
    lines(chains[[c]][, param_index], col = cols[c])
  }
}
```


# Spike-and-slab summaries

The spike-and-slab prior yields posterior inclusion probabilities for
edges:

```{r}
coef(fit)$indicator
```
- Values near 1.0: strong evidence the edge is present.
- Values near 0.0: strong evidence the edge is absent.
- Values near 0.5: inconclusive (absence of evidence).

# Bayes factors

When the prior inclusion probability for an edge is equal to 0.5 (e.g., using
`edge_prior = bernoulli_prior(0.5)`, the default, or a symmetric Beta prior
`edge_prior = beta_bernoulli_prior(alpha, beta)` with `alpha == beta`), we can
directly transform inclusion probabilities into
Bayes factors for edge presence vs absence:

```{r}
# Example for one edge
p = coef(fit)$indicator[1, 5]
BF_10 = p / (1 - p)
BF_10
```

Here the Bayes factor in favor of inclusion (H1) is small, meaning that there is little evidence for inclusion. Since the Bayes factor is transitive, we can 
use it to express the evidence in favor of exclusion (H0) as

```{r}
1 / BF_10
```
This Bayes factor shows that there is strong evidence for the absence of a 
network relation between the variables `intrusion` and `physior`.

# NUTS diagnostics

When using `update_method = "nuts"` (the default), additional diagnostics are 
available to assess the quality of the Hamiltonian Monte Carlo sampling. These
can be accessed via `fit$nuts_diag`:

```{r}
fit$nuts_diag$summary
```

## E-BFMI

E-BFMI (Energy Bayesian Fraction of Missing Information) measures how  efficiently the sampler explores the posterior. It compares the typical size of energy changes between successive samples to the overall spread of energies. Values close to 1 indicate that the sampler moves freely across the energy landscape; values below 0.3 suggest the sampler may be getting stuck or that the chain has not yet settled into its stationary distribution.

A low E-BFMI does not necessarily mean your results are wrong, but it does warrant further investigation. In models with edge selection, the most common cause is that the warmup period was too short for the discrete graph structure to equilibrate. Increasing `warmup` often resolves this.

## Divergent transitions

Divergent transitions occur when the numerical integrator encounters regions of the posterior where the curvature changes too rapidly for the current step size. A small number of divergences (say, fewer than 0.1% of samples) is generally acceptable. However, many divergences indicate that the sampler may be missing important parts of the posterior.

If you see a large number of divergences, consider increasing `target_accept` (which makes the sampler use a smaller step size) and, if this does not fix it, switching to `update_method = "adaptive-metropolis"`.

## Tree depth

NUTS builds trajectories by repeatedly doubling their length until a "U-turn" criterion is satisfied. If the trajectory frequently reaches the maximum allowed depth (`nuts_max_depth`, default 10), it suggests the sampler may benefit from longer trajectories to explore the posterior efficiently. Hitting the maximum depth occasionally is normal; hitting it on most iterations may indicate challenging posterior geometry. If this happens, consider increasing `nuts_max_depth`.

## Non-reversible steps

For MRFs with continuous variables, the leapfrog integrator enforces equality constraints through a projection step. After each forward step, the integrator checks whether reversing the step returns to the starting point. When the round-trip error exceeds a tolerance scaled by the square of the step size, the step is flagged as non-reversible.

A small number of non-reversible steps is not a concern. A large number indicates that the step size is too large for the constraint geometry. Because the step size is tuned during warmup, the most effective remedy is to increase `warmup` so the adapter has more time to find an appropriate step size. If non-reversible steps persist after increasing warmup, switch to `update_method = "adaptive-metropolis"`.

## Warmup and equilibration

Standard HMC/NUTS warmup is designed to tune the step size and mass matrix for the continuous parameters. In models with edge selection, the discrete graph structure may take longer to reach its stationary distribution than the continuous parameters. As a result, even after warmup completes, the first portion of the sampling phase may still show transient behavior (i.e., non-stationarity).
  
The `warmup_check` component provides simple diagnostics that compare the first and second halves of the post-warmup samples:

```{r}
fit$nuts_diag$warmup_check
```

The returned list contains the following fields (one value per chain):

- **warmup_incomplete**: A logical flag that is `TRUE` when any of the indicators below suggest the chain may not have reached stationarity.
- **energy_slope**: The slope of a linear regression of energy against iteration number. A slope near zero indicates stable energy; a significant negative slope suggests the chain is still drifting toward higher-probability regions.
- **slope_significant**: `TRUE` if the energy slope is statistically significant (p < 0.01).
- **ebfmi_first_half** and **ebfmi_second_half**: E-BFMI computed separately for the first and second halves of the post-warmup samples. If the first-half value is much lower (for example, below 0.3) while the second-half value is healthy, the early samples were likely still settling.
- **var_ratio**: The ratio of energy variance in the first half to that in the second half. A ratio much greater than 1 (for example, above 2) indicates higher variability early on, consistent with transient behavior.

If these diagnostics suggest the chain was still settling, increase `warmup` and re-run the model. If diagnostics remain problematic after a substantial increase (for example, doubling or tripling `warmup`), consider re-fitting with `update_method = "adaptive-metropolis"` and comparing the posterior summaries. If the two samplers produce similar results, the estimates are likely trustworthy despite the warnings; if they differ substantially, that warrants further investigation of the model or data.

# Z-ratio trust gauge

The hierarchical graph prior (`precision_graph_prior = "hierarchical"`) accepts edge moves using an approximate normalizing-constant ratio. The trust gauge measures how often that approximation would have changed an accept/reject decision, by comparing it against a block-local Monte Carlo reference on the chain's own moves. It flags; it does not fix. The summary is returned as `fit$zratio_diag` and is available through `summarize_zratio_gauge()`.

## D: the operational flag

`D` is the fraction of edge moves that would flip under the exact ratio. A chain is green when `D` stays under 1% (beyond the reference's own Monte Carlo noise floor) and amber above it. Unlike a *p*-value, `D` is in direct impact units, so the same 1% tolerance applies at any number of variables and any graph prior.

If a chain is amber, the approximation is materially affecting that chain's structure decisions in the regime it explored (typically a high prior scale and/or many variables). Options: reduce the prior scale toward the calibrated range, or treat the recovered structure as approximate and cross-check the flagged edges.

## Two-tier reading: D and s_e

The gauge reports both `D` and the log-ratio error `s_e`, read together the way E-BFMI and R-hat are read together:

- `s_e` (in nats) answers *is the estimator biased at all?* It is sensitive and fires whenever the approximation is off, even when it changes no decisions.
- `D` answers *does that bias change this chain's moves?* It is the operational flag. With strong data, `D` can be near zero even when `s_e` is non-zero, because the likelihood saturates the acceptances and the biased ratio flips nothing (correct self-calibration). So watch `D` for the flag; read `s_e` when you want to know whether the estimator itself is off (for example, a prior-only or weak-data run).

## Relationship to R-hat

R-hat catches mixing; the gauge catches estimator fidelity. They usually co-fire, but if every chain shares the same systematic ratio bias and still mixes, the chains agree and R-hat looks clean — the gauge is what closes that blind spot. Read them together.

# Next steps

- See *Getting Started* for a simple one-sample workflow.
- See *Model Comparison* for group differences.
