---
title: "Speed and Scale in Gaussian Graphical Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speed and Scale in Gaussian Graphical Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2
)
library(bgms)
# Every number in this vignette comes from this object. It was produced by one
# run of data-raw/make-fast-ggm.R, whose calls are the calls shown below.
B = readRDS("fast-ggm.rds")
sec = function(x) sprintf("%.1f", x)
# knitr's inline hook renders counts of this size in scientific notation.
num = function(x) formatC(x, format = "d", big.mark = ",")
```

# When runtime is the constraint

Gaussian graphical models often contain hundreds of variables, and edge
selection is computationally demanding at that size: a graph with two hundred
variables has 19,900 candidate edges, and the sampler updates the graph and
its precision matrix over and over. For such analyses `bgm()` provides a
faster route, selected by two arguments:

```{r, eval = FALSE}
fit = bgm(y,
  variable_type = "continuous",
  precision_graph_prior = "joint",
  update_method = "gibbs"
)
```

`update_method = "gibbs"` replaces the default No-U-Turn sampler with a
conjugate sampler that redraws an entire column of the precision matrix at a
time. `precision_graph_prior = "joint"` normalizes the prior over the precision
matrix and graph once, globally, rather than separately for each graph. This
puts every term in the edge move in closed form. Neither option is the default,
and the second changes the model being fitted. That change is described first
below; the timings follow.

On an Apple M5 Pro with four chains on four threads, the two options together
speed up a fifty-variable fit by a factor of
`r sprintf("%.0f", B$mid$speedup)`, and they complete a two-hundred-variable
fit (19,900 candidate edges) in about
`r sprintf("%.0f", B$high$elapsed / 60)` minutes.


# The trade-off

The two routes do not target the same model, and the difference is in the prior
over graphs.

`bgm()`'s default, `precision_graph_prior = "hierarchical"`, normalizes the
prior over the precision matrix separately within each graph. Integrating the
precision matrix out then returns the stated edge prior exactly: set
`bernoulli_prior(0.5)` and every edge really does carry prior probability
one half. The price is that each edge move has to evaluate a graph-dependent
normalizing constant.

The joint specification normalizes once for the whole model. That is what makes
the edge move cheap (the constant is the same on both sides of the comparison
and cancels), but it also means the edge prior is no longer the prior over
graphs. What survives is the edge prior reweighted by each graph's normalizing
constant, and that reweighting favors sparser graphs. On six variables, with a
nominal inclusion probability of one half:

```{r, eval = FALSE}
extract_prior_inclusion_probabilities(fit)[1, 2]
```

```{r, echo = FALSE}
data.frame(
  route = c("hierarchical (default)", "joint"),
  nominal = c(B$prior$nominal, B$prior$nominal),
  realized = round(c(B$prior$hierarchical, B$prior$joint), 3)
)
```

The default returns the stated half. The joint specification returns
`r round(B$prior$joint, 3)`, and the gap widens as the model grows. Nothing here
is broken (the joint specification is a coherent model, and the sampler targets
its posterior correctly), but it is a *different* model, so its
inclusion Bayes factors are not interchangeable with the default's. An edge that
is undecided on one route can be evidence of absence on the other, without
either fit being wrong.

`bgm()` reports this when it fits:

```
Joint precision-graph specification: the realized edge-inclusion prior is the
edge prior reweighted by the per-graph normalizer, not the nominal edge prior
```

The choice between specifications is a modelling choice, not a tuning setting.
Use the fast route when runtime is the obstacle, and report which specification
produced the results. An analysis that reports a stated prior inclusion
probability needs the nominal edge prior to be the prior over graphs, which
only the default specification provides.

One practical consequence. Reading the realized prior back, which
`extract_prior_inclusion_probabilities()`, `extract_inclusion_bf()` and
`verdicts()` all need, requires a table built by simulation for the model size
at hand. The table is computed once and cached on disk, but it is not quick:
at six variables it takes a few seconds, and at fifty it takes longer than the
fit itself. Posterior inclusion probabilities, from `summary()` or
`extract_posterior_inclusion_probabilities()`, need nothing of the kind.


# A mid-sized model, both ways

The first example has `r B$mid$p` variables and `r B$mid$n` observations,
drawn from a sparse Gaussian graphical model with `r B$mid$edges` edges among
`r B$mid$pairs` pairs. Both fits use the package defaults for run length
(`r formals(bgm)$warmup` warmup and `r formals(bgm)$iter` retained iterations)
and both run four chains on four threads. They differ in nothing else.

```{r, eval = FALSE}
default = bgm(y,
  variable_type = "continuous",
  chains = 4, cores = 4, seed = 2026
)

fast = bgm(y,
  variable_type = "continuous",
  precision_graph_prior = "joint",
  update_method = "gibbs",
  chains = 4, cores = 4, seed = 2026
)
```

```{r, echo = FALSE}
data.frame(
  route = c("hierarchical + nuts (default)", "joint + gibbs"),
  seconds = c(round(B$mid$default$elapsed, 1), round(B$mid$fast$elapsed, 1)),
  edges_found = c(B$mid$default$selected, B$mid$fast$selected),
  true_edges_found = c(B$mid$default$detected, B$mid$fast$detected),
  false_positives = c(
    B$mid$default$false_positive, B$mid$fast$false_positive
  ),
  max_rhat = round(c(B$mid$default$max_rhat, B$mid$fast$max_rhat), 3)
)
```

The default specification takes `r sec(B$mid$default$elapsed)` seconds and the
fast specification `r sec(B$mid$fast$elapsed)`, a factor of
`r sprintf("%.0f", B$mid$speedup)`. Both recovered `r B$mid$fast$detected` of
the `r B$mid$edges` true edges at the conventional 0.5 cut, and both are
converged.

The two edge sets agree on `r B$mid$agreement` of the `r B$mid$pairs` pairs.
That agreement is an observation about this data set, not a guarantee: the
sparser realized prior of the joint specification pulls verdicts toward
absence, and how far depends on how informative the data are. On data with
weaker signal the two specifications will diverge sooner. The message of the
table is the seconds column: the recovery is the same, the runtime is not.


# Two hundred variables

The second example has `r B$high$p` variables and `r B$high$n` observations:
`r num(B$high$pairs)` candidate edges, `r B$high$edges` of them real. Only the
fast route was run.

```{r, eval = FALSE}
fit = bgm(y,
  variable_type = "continuous",
  precision_graph_prior = "joint",
  update_method = "gibbs",
  chains = 4, cores = 4, seed = 2026
)
```

```{r, echo = FALSE}
data.frame(
  variables = B$high$p,
  candidate_edges = B$high$pairs,
  seconds = round(B$high$elapsed, 1),
  edges_found = B$high$selected,
  true_edges_found = B$high$detected,
  false_positives = B$high$false_positive,
  max_rhat = round(B$high$max_rhat, 3)
)
```

The fit takes `r sec(B$high$elapsed)` seconds (about
`r sprintf("%.0f", B$high$elapsed / 60)` minutes) for four chains on four
threads at the package's default run length, deciding `r num(B$high$pairs)`
edges. It found `r B$high$detected` of the `r B$high$edges` true edges and
`r B$high$false_positive` pairs that are not edges, which on the
`r num(B$high$pairs - B$high$edges)` non-edges is a false-positive rate of
`r sprintf("%.1f%%", 100 * B$high$false_positive / (B$high$pairs - B$high$edges))`.

Nothing about reading the fit changes at this size:

```{r, eval = FALSE}
head(summary(fit)$indicator)
```

```{r, echo = FALSE}
round(B$high$head, 3)
```

The remaining question, across all `r num(B$high$pairs)` pairs, is how decided
the posterior is. The middle of the range holds `r B$high$profile$pairs[3]`
pairs
(`r sprintf("%.1f%%", 100 * B$high$profile$pairs[3] / B$high$pairs)` of the
candidates), so the pairs that need a closer look form a short list:

```{r, echo = FALSE}
knitr::kable(B$high$profile, col.names = c("inclusion probability", "pairs"))
```

The default route was not run at this size, so this vignette makes no claim
about what it would have cost. The measured comparison is the one at fifty
variables above.


# How it scales

Same generator, same `r B$high$n` observations, same four chains on four
threads, the fast route throughout.

```{r, echo = FALSE}
knitr::kable(
  data.frame(
    variables = B$scaling$p,
    candidate_edges = B$scaling$pairs,
    seconds = round(B$scaling$elapsed, 1)
  )
)
```

```{r, echo = FALSE, fig.alt = "Wall-clock seconds against the number of variables, on a logarithmic vertical axis, rising steeply from a fraction of a second at 25 variables to several minutes at 200."}
op = par(mar = c(4.2, 4.4, 1.2, 1.2), las = 1, bty = "n", cex = 0.9)
plot(B$scaling$p, B$scaling$elapsed,
  log = "y", type = "b", pch = 19, lwd = 2, cex = 1.1,
  col = "#2f6f9f",
  xlab = "Number of variables",
  ylab = "Seconds (four chains, four threads)",
  panel.first = grid(col = "grey92", lty = 1, lwd = 1)
)
text(B$scaling$p, B$scaling$elapsed,
  labels = sprintf("%.1f", B$scaling$elapsed),
  pos = c(4, 4, 4, 4, 2), offset = 0.6, cex = 0.8, col = "grey30"
)
par(op)
```

What drives that curve is the number of variables, not the number of
observations. Complete data enter the sampler only through the cross-product of
the data matrix, formed once before sampling starts; every draw after that
works on the precision matrix, whose size the variables alone set. More
observations change what the sampler is drawing from, not how much work a draw
costs. More variables change both.


# Where the fast route applies

`update_method = "gibbs"` applies only to the Gaussian graphical model.
Ordinal, Blume-Capel or mixed data are refused with an error:

```{r, error = TRUE}
bgm(Wenchuan[, 1:4], update_method = "gibbs")
```

Within continuous data it needs conjugate ingredients: a `normal_prior()` or
`cauchy_prior()` on the interactions, and `exponential_prior()` or
`gamma_prior()` on the precision diagonal. Those are the defaults for
`interaction_prior` and `precision_scale_prior`, so the usual call already
qualifies; `beta_prime_prior()` on the interactions does not. Edge selection is
supported for both slabs, and so is `edge_selection = FALSE`, which estimates
the precision matrix at a fixed graph.

`precision_graph_prior = "joint"` is separate from the sampler and applies to
continuous data under edge selection, in Gaussian graphical models and in the
continuous block of a mixed model. The two arguments compose but do not depend
on each other: the Gibbs sampler runs under the default hierarchical
specification, and the joint specification runs under `"nuts"`.

One combination is fast to sample but slow to set up. Under the joint
specification a learned inclusion probability (`beta_bernoulli_prior()` or
`sbm_prior()`) needs the same simulated correction table described above,
because the constant that cancels for a fixed inclusion probability stops
cancelling once that probability is itself being updated. The table is built at
the first fit of a model configuration and cached, but at large numbers of
variables the wait for the table far exceeds the sampling time. For a fast run
at scale, keep `bernoulli_prior()`, the default.


# Background

The prior used here specifies the precision matrix entry by entry: an edge that
is absent has its entry fixed at exactly zero, and an edge that is present draws
its entry from a diffuse slab. The construction comes from the Bayesian
variable-selection literature [@george1993jasa_variable] and was brought to
Gaussian graphical models by @wang2015ba_scaling, whose column-wise block Gibbs
sampler is the ancestor of the one `update_method = "gibbs"` runs. One
divergence from that ancestor matters for runtime: Wang's spike is continuous,
a narrow normal around zero, so every column update redraws all of the
column's entries. `bgms` uses a point mass, so a column update only works
through the entries the current graph includes, and a sparser graph makes a
cheaper update. The appeal of the construction is that it avoids a
graph-dependent normalizing constant. The
G-Wishart prior is the literature's canonical case of that cost
[@atay-kayis2005biometrika_monte; @mohammadi2023jasa_accelerating], but the
cost is not particular to the G-Wishart: `bgm()`'s own hierarchical
specification pays a constant of the same kind, deliberately, so that the
stated edge prior is exactly the prior over graphs. The joint specification
takes the classical avoidance and the hierarchical specification declines it;
that is the choice `precision_graph_prior` exposes.
@vogels2024jasa_bayesian survey the wider field and compare the available
samplers empirically.

`bgms` adds one thing to the classical construction. Specifying the precision
matrix entry by entry does not guarantee that the result is positive definite,
so the prior has to be truncated to the cone of positive-definite matrices, and
that truncation distorts it more and more as the number of variables grows
[@jewson2024biometrics_graphical]. `bgm()` counters this by tilting the prior by
a power `delta` of the determinant, which pushes mass away from the boundary of
the cone; `delta` defaults to half the logarithm of the number of variables, so
larger models are tilted harder. This is why the numbers above hold up at two
hundred variables rather than collapsing onto the empty graph.

Scaling this prior is an active problem and `bgms` is not alone on it.
@sulem2025arxiv_bayesian give the closest treatment:
Metropolis-Hastings within block Gibbs on the same spike-and-slab Gaussian
graphical model, with mixing bounds that are dimension-free under suitable
conditions and a global proposal that adds or removes several edges in one
move. They report carrying exact Bayesian inference from roughly one hundred
variables to roughly one thousand. Their work is implemented in the CRAN package
**modelSelection** (version 1.0.7, 2026-05-16), which covers Bayesian model
selection and averaging for regression, generalized linear and additive models,
mixtures and graphical models. **ssgraph** (version 1.16, 2025-08-29) is the
other near neighbor, doing Bayesian structure learning for undirected graphical
models with spike-and-slab priors on continuous, discrete and mixed data.

How does the fast route compare with these two packages? A measured
comparison is maintained on the package website's
[comparison page](https://bayesian-graphical-modelling-lab.github.io/bgms/guide/fast-ggm.html),
with the full protocol, the seeded scripts, and the complete tables
[in its repository](https://github.com/Bayesian-Graphical-Modelling-Lab/bgms-docs/tree/main/benchmarks/ggm-comparison).
It lives there rather than here because its results depend on the competitor
versions of the day, which change on a faster clock than package releases. In
summary: because default run lengths differ by orders of magnitude across the
packages, the comparison measures the time to a converged edge ranking, the
convergence-cost metric of @vogels2024jasa_bayesian. On replicate datasets
from this vignette's generator, `bgms` reaches its converged ranking in
seconds at every tested sample size, a median 18 to 25 times sooner than
ssgraph under a matched edge prior. modelSelection converges within about a
second, so its short default run is converged rather than undersampled; it
targets a sparser prior, recovers somewhat fewer edges at a matched threshold,
and cannot be seeded. The three packages arrive at nearly identical edge
rankings; they differ in how quickly they get there and in where their
default priors place the selection cut.

Readers whose problem is a single very large Gaussian graphical model and
nothing else should look at all three; what `bgms` adds around this route is
the rest of the package: ordinal, Blume-Capel and mixed models, group
comparison, prior sensitivity, and the determinant tilt that keeps the prior
usable as the number of variables grows.


# Reproducing the timings

Every number above was produced by `data-raw/make-fast-ggm.R`, which is in the
package repository and shows the generator, the seeds and the calls. Each fit
ran on its own, four chains on four threads.

```{r, echo = FALSE}
data.frame(
  field = c("CPU", "cores available", "threads used", "chains",
            "operating system", "R", "bgms", "run"),
  value = c(B$machine$cpu, B$machine$cores_total, B$machine$threads,
            B$machine$chains, B$machine$os, B$machine$r_version,
            B$machine$bgms_version, B$machine$date)
)
```

The timings are specific to this machine and its thread counts; the ratios
should transfer better than the absolute numbers.


# References
