---
title: "Speed and Scale in Gaussian Graphical Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speed and Scale in Gaussian Graphical Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2
)
library(bgms)
# Every number in this vignette comes from this object. It was produced by one
# run of data-raw/make-fast-ggm.R, whose calls are the calls shown below.
B = readRDS("fast-ggm.rds")
sec = function(x) sprintf("%.1f", x)
# knitr's inline hook renders counts of this size in scientific notation.
num = function(x) formatC(x, format = "d", big.mark = ",")
```

# When runtime is the constraint

Continuous data give a Gaussian graphical model, and continuous data are where
graphs get large. Two hundred variables is twenty thousand candidate edges, and
an analysis that decides them one by one is a long analysis. For that case
`bgm()` has a faster route. Two arguments select it:

```{r, eval = FALSE}
fit = bgm(y,
  variable_type = "continuous",
  precision_graph_prior = "joint",
  update_method = "gibbs"
)
```

`update_method = "gibbs"` replaces the default No-U-Turn sampler with a
conjugate sampler that redraws a whole column of the precision matrix at a time.
`precision_graph_prior = "joint"` normalizes the prior over the precision matrix
and the graph once, globally, instead of once per graph, which puts every term
of the edge move in closed form. Neither is the default, and the second one
changes the model being fitted. That change comes first below; the timings
follow.

On an Apple M5 Pro, four chains on four threads, at fifty variables the pair is
worth a factor of `r sprintf("%.0f", B$mid$speedup)`. At two hundred variables
(twenty thousand candidate edges) it decides all of them in
`r sprintf("%.0f", B$high$elapsed / 60)` minutes.


# The trade-off

The two routes do not target the same model, and the difference is in the prior
over graphs.

`bgm()`'s default, `precision_graph_prior = "hierarchical"`, normalizes the
prior over the precision matrix separately within each graph. Integrating the
precision matrix out then returns the stated edge prior exactly: set
`bernoulli_prior(0.5)` and every edge really does carry prior probability
one half. The price is that each edge move has to evaluate a graph-dependent
normalizing constant.

The joint specification normalizes once for the whole model. That is what makes
the edge move cheap (the constant is the same on both sides of the comparison
and cancels), but it also means the edge prior is no longer the prior over
graphs. What survives is the edge prior reweighted by each graph's normalizing
constant, and that reweighting favors sparser graphs. On six variables, with a
nominal inclusion probability of one half:

```{r, eval = FALSE}
extract_prior_inclusion_probabilities(fit)[1, 2]
```

```{r, echo = FALSE}
data.frame(
  route = c("hierarchical (default)", "joint"),
  nominal = c(B$prior$nominal, B$prior$nominal),
  realized = round(c(B$prior$hierarchical, B$prior$joint), 3)
)
```

The default returns the stated half. The joint specification returns
`r round(B$prior$joint, 3)`, and the gap widens as the model grows. Nothing here
is broken (the joint specification is a coherent model, and the sampler targets
its posterior correctly), but it is a *different* model, so its
inclusion Bayes factors are not interchangeable with the default's. An edge that
is undecided on one route can be evidence of absence on the other, without
either fit being wrong.

`bgm()` says so when it fits:

```
Joint precision-graph specification: the realized edge-inclusion prior is the
edge prior reweighted by the per-graph normalizer, not the nominal edge prior
```

The choice is a modelling choice, not a tuning knob. Take the fast route when
the size of the problem is the obstacle, and report which specification the
answer came from. An analysis whose report includes a stated prior inclusion
probability needs the nominal edge prior to be the prior over graphs, and that
is the default route.

One practical consequence. Reading the realized prior back, which
`extract_prior_inclusion_probabilities()`, `extract_inclusion_bf()` and
`verdicts()` all need, requires a table built by simulation for the model size
at hand. It is computed once and cached on disk, but it is not quick: at six
variables it is a few seconds, and by fifty it is longer than the fit it
describes. Posterior inclusion probabilities, from `summary()` or
`extract_posterior_inclusion_probabilities()`, need nothing of the kind.


# A mid-sized model, both ways

`r B$mid$p` variables, `r B$mid$n` observations, drawn from a sparse Gaussian
graphical model with `r B$mid$edges` edges among `r B$mid$pairs` pairs. Both
fits use the package defaults for length (`r formals(bgm)$warmup` warmup and
`r formals(bgm)$iter` retained iterations) and both run four chains on four
threads. They differ in nothing else.

```{r, eval = FALSE}
default = bgm(y,
  variable_type = "continuous",
  chains = 4, cores = 4, seed = 2026
)

fast = bgm(y,
  variable_type = "continuous",
  precision_graph_prior = "joint",
  update_method = "gibbs",
  chains = 4, cores = 4, seed = 2026
)
```

```{r, echo = FALSE}
data.frame(
  route = c("hierarchical + nuts (default)", "joint + gibbs"),
  seconds = c(round(B$mid$default$elapsed, 1), round(B$mid$fast$elapsed, 1)),
  edges_found = c(B$mid$default$selected, B$mid$fast$selected),
  true_edges_found = c(B$mid$default$detected, B$mid$fast$detected),
  false_positives = c(
    B$mid$default$false_positive, B$mid$fast$false_positive
  ),
  max_rhat = round(c(B$mid$default$max_rhat, B$mid$fast$max_rhat), 3)
)
```

`r sec(B$mid$default$elapsed)` seconds against `r sec(B$mid$fast$elapsed)`, a
factor of `r sprintf("%.0f", B$mid$speedup)`, on four chains on four threads.
Both routes recovered `r B$mid$fast$detected` of the `r B$mid$edges` true edges
at the conventional half cut, and both are converged.

The two edge sets agree on `r B$mid$agreement` of the `r B$mid$pairs` pairs.
That agreement is an observation about this data set, not a guarantee: the
sparser realized prior of the joint specification will pull verdicts toward
absence, and how far it pulls them depends on how much the data have to say. On
data with weaker signal the two routes will part company sooner. The point of
the table is the left-hand column.


# Two hundred variables

`r B$high$p` variables and `r B$high$n` observations: `r num(B$high$pairs)`
candidate edges, `r B$high$edges` of them real. The fast route only.

```{r, eval = FALSE}
fit = bgm(y,
  variable_type = "continuous",
  precision_graph_prior = "joint",
  update_method = "gibbs",
  chains = 4, cores = 4, seed = 2026
)
```

```{r, echo = FALSE}
data.frame(
  variables = B$high$p,
  candidate_edges = B$high$pairs,
  seconds = round(B$high$elapsed, 1),
  edges_found = B$high$selected,
  true_edges_found = B$high$detected,
  false_positives = B$high$false_positive,
  max_rhat = round(B$high$max_rhat, 3)
)
```

`r sec(B$high$elapsed)` seconds (about
`r sprintf("%.0f", B$high$elapsed / 60)` minutes) for four chains on four
threads at the package's default length, deciding `r num(B$high$pairs)` edges.
It found `r B$high$detected` of the `r B$high$edges` true edges and
`r B$high$false_positive` pairs that are not edges, which on the
`r num(B$high$pairs - B$high$edges)` non-edges is a false-positive rate of
`r sprintf("%.1f%%", 100 * B$high$false_positive / (B$high$pairs - B$high$edges))`.

Nothing about reading the fit changes at this size:

```{r, eval = FALSE}
head(summary(fit)$indicator)
```

```{r, echo = FALSE}
round(B$high$head, 3)
```

The remaining question, across all `r num(B$high$pairs)` pairs, is how decided
the posterior is. The middle of the range holds `r B$high$profile$pairs[3]`
pairs
(`r sprintf("%.1f%%", 100 * B$high$profile$pairs[3] / B$high$pairs)` of the
candidates), so the pairs that need a second look are a short list rather
than a second analysis:

```{r, echo = FALSE}
knitr::kable(B$high$profile, col.names = c("inclusion probability", "pairs"))
```

The default route was not run at this size, so this vignette makes no claim
about what it would have cost. The measured comparison is the one at fifty
variables above.


# How it scales

Same generator, same `r B$high$n` observations, same four chains on four
threads, the fast route throughout.

```{r, echo = FALSE}
knitr::kable(
  data.frame(
    variables = B$scaling$p,
    candidate_edges = B$scaling$pairs,
    seconds = round(B$scaling$elapsed, 1)
  )
)
```

```{r, echo = FALSE, fig.alt = "Wall-clock seconds against the number of variables, on a logarithmic vertical axis, rising steeply from a fraction of a second at 25 variables to several minutes at 200."}
op = par(mar = c(4.2, 4.4, 1.2, 1.2), las = 1, bty = "n", cex = 0.9)
plot(B$scaling$p, B$scaling$elapsed,
  log = "y", type = "b", pch = 19, lwd = 2, cex = 1.1,
  col = "#2f6f9f",
  xlab = "Number of variables",
  ylab = "Seconds (four chains, four threads)",
  panel.first = grid(col = "grey92", lty = 1, lwd = 1)
)
text(B$scaling$p, B$scaling$elapsed,
  labels = sprintf("%.1f", B$scaling$elapsed),
  pos = c(4, 4, 4, 4, 2), offset = 0.6, cex = 0.8, col = "grey30"
)
par(op)
```

What drives that curve is the number of variables, not the number of
observations. Complete data enter the sampler only through the cross-product of
the data matrix, formed once before sampling starts; every draw after that
works on the precision matrix, whose size the variables alone set. More
observations change what the sampler is drawing from, not how much work a draw
costs. More variables change both.


# Where the fast route applies

`update_method = "gibbs"` is the Gaussian graphical model's sampler and nothing
else. Ordinal, Blume-Capel or mixed data are refused outright:

```{r, error = TRUE}
bgm(Wenchuan[, 1:4], update_method = "gibbs")
```

Within continuous data it needs conjugate ingredients: a `normal_prior()` or
`cauchy_prior()` on the interactions, and `exponential_prior()` or
`gamma_prior()` on the precision diagonal. Those are the defaults for
`interaction_prior` and `precision_scale_prior`, so the usual call already
qualifies; `beta_prime_prior()` on the interactions does not. Edge selection is
supported for both slabs, and so is `edge_selection = FALSE`, which estimates
the precision matrix at a fixed graph.

`precision_graph_prior = "joint"` is separate from the sampler and applies to
continuous data under edge selection, in Gaussian graphical models and in the
continuous block of a mixed model. The two arguments compose but do not depend
on each other: the Gibbs sampler runs under the default hierarchical
specification, and the joint specification runs under `"nuts"`.

One combination is fast to fit but slow to start. Under the joint
specification a learned inclusion probability (`beta_bernoulli_prior()` or
`sbm_prior()`) needs the same simulated correction table described above,
because the constant that cancels for a fixed inclusion probability stops
cancelling once that probability is itself being updated. The table is built at
the first fit of a model configuration and cached, but at large numbers of
variables that first fit will be waiting on it for far longer than the sampling
takes. For a fast run at scale, keep `bernoulli_prior()`, the default.


# Background

The prior used here specifies the precision matrix entry by entry: an edge that
is absent has its entry fixed at exactly zero, and an edge that is present draws
its entry from a diffuse slab. The construction comes from the Bayesian
variable-selection literature [@george1993jasa_variable] and was brought to
Gaussian graphical models by @wang2015ba_scaling, whose column-wise block Gibbs
sampler is the ancestor of the one `update_method = "gibbs"` runs. Its appeal
was, and is, that it avoids the graph-dependent normalizing constant that makes
structure learning expensive under the G-Wishart prior
[@atay-kayis2005biometrika_monte; @mohammadi2023jasa_accelerating]. That
avoidance is exactly what the joint specification does and what the
hierarchical specification declines to do, which is the choice
`precision_graph_prior` exposes. @vogels2024jasa_bayesian survey the wider field
and compare the available samplers empirically.

`bgms` adds one thing to the classical construction. Specifying the precision
matrix entry by entry does not guarantee that the result is positive definite,
so the prior has to be truncated to the cone of positive-definite matrices, and
that truncation distorts it more and more as the number of variables grows
[@jewson2024biometrics_graphical]. `bgm()` counters this by tilting the prior by
a power `delta` of the determinant, which pushes mass away from the boundary of
the cone; `delta` defaults to half the logarithm of the number of variables, so
larger models are tilted harder. This is why the numbers above hold up at two
hundred variables rather than collapsing onto the empty graph.

Scaling this prior is an active problem and `bgms` is not alone on it.
@sulem2025arxiv_bayesian give the closest treatment:
Metropolis-Hastings within block Gibbs on the same spike-and-slab Gaussian
graphical model, with mixing bounds that are dimension-free under suitable
conditions and a global proposal that adds or removes several edges in one
move. They report carrying exact Bayesian inference from roughly one hundred
variables to roughly one thousand. Their work is implemented in the CRAN package
**modelSelection** (version 1.0.7, 2026-05-16), which covers Bayesian model
selection and averaging for regression, generalized linear and additive models,
mixtures and graphical models. **ssgraph** (version 1.16, 2025-08-29) is the
other near neighbor, doing Bayesian structure learning for undirected graphical
models with spike-and-slab priors on continuous, discrete and mixed data.

Neither package was run against `bgms` on the same data for this vignette, so it
makes no comparative speed claim about either. Every timing above is `bgms`
against `bgms`. Readers whose problem is a single very large Gaussian graphical
model and nothing else should look at all three; what `bgms` adds around this
route is the rest of the package: ordinal, Blume-Capel and mixed models,
group comparison, prior sensitivity, and the determinant tilt that keeps the
prior usable as the number of variables grows.


# Reproducing the timings

Every number above was produced by `data-raw/make-fast-ggm.R`, which is in the
package repository and shows the generator, the seeds and the calls. Each fit
ran on its own, four chains on four threads.

```{r, echo = FALSE}
data.frame(
  field = c("CPU", "cores available", "threads used", "chains",
            "operating system", "R", "bgms", "run"),
  value = c(B$machine$cpu, B$machine$cores_total, B$machine$threads,
            B$machine$chains, B$machine$os, B$machine$r_version,
            B$machine$bgms_version, B$machine$date)
)
```

Timings are facts about a machine. The ratios will travel further than the
seconds.


# References
