---
title: "Speed and Scale in Gaussian Graphical Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speed and Scale in Gaussian Graphical Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2
)
library(bgms)
# The worked bgms results in this vignette come from this object. It was
# produced by one run of data-raw/make-fast-ggm.R, whose calls are shown below.
# The cross-package comparison summarized later is maintained separately.
B = readRDS("fast-ggm.rds")
sec = function(x) sprintf("%.1f", x)
# knitr's inline hook renders counts of this size in scientific notation.
num = function(x) formatC(x, format = "d", big.mark = ",")
```

# When runtime is the constraint

Gaussian graphical models often contain hundreds of variables, and edge
selection is computationally demanding at that size: a graph with two hundred
variables has 19,900 candidate edges, and the sampler repeatedly updates the
graph and its precision matrix. For such analyses, `bgm()` provides a faster
route, selected by two arguments:

```{r, eval = FALSE}
fit = bgm(y,
  variable_type = "continuous",
  precision_graph_prior = "joint",
  update_method = "gibbs"
)
```

`update_method = "gibbs"` replaces the default No-U-Turn sampler with a
conjugate sampler that redraws an entire column of the precision matrix at a
time. `precision_graph_prior = "joint"` normalizes the prior over the precision
matrix and graph once, globally, rather than separately for each graph. This
puts every term in the edge move in closed form. Neither option is the default,
and the second changes the model being fitted. That change is described first
below; the timings follow.

On an Apple M5 Pro with four chains on four threads, this combination reduces
the runtime in the fifty-variable example by a factor of
`r sprintf("%.0f", B$mid$speedup)` and completes the two-hundred-variable
example, with 19,900 candidate edges, in about
`r sprintf("%.0f", B$high$elapsed / 60)` minutes.


# The trade-off

The two routes do not target the same model, and the difference is in the prior
over graphs.

`bgm()`'s default, `precision_graph_prior = "hierarchical"`, normalizes the
prior over the precision matrix separately within each graph. Integrating the
precision matrix out then returns the stated edge prior exactly: set
`bernoulli_prior(0.5)` and every edge has prior probability one half. The price
is that each edge move must evaluate a graph-dependent normalizing constant.

The joint specification normalizes once for the whole model. That is what makes
the edge move cheap: the constant is the same on both sides of the comparison
and cancels. But it also means that the edge prior is no longer the prior over
graphs. What survives is the edge prior reweighted by each graph's normalizing
constant, and that reweighting favors sparser graphs. On six variables, with a
nominal inclusion probability of one half:

```{r, eval = FALSE}
extract_prior_inclusion_probabilities(fit)[1, 2]
```

```{r, echo = FALSE}
data.frame(
  route = c("hierarchical (default)", "joint"),
  nominal = c(B$prior$nominal, B$prior$nominal),
  realized = round(c(B$prior$hierarchical, B$prior$joint), 3)
)
```

The default returns the stated half. The joint specification returns
`r round(B$prior$joint, 3)`, and the gap widens as the model grows. Nothing here
is broken: the joint specification is a coherent model, and the sampler targets
its posterior correctly. But it is a *different* model, so its inclusion Bayes
factors are not interchangeable with the default's. An edge that is undecided
under one specification can show evidence of absence under the other, without
either fit being wrong.

`bgm()` reports this when it fits:

```
Joint precision-graph specification: the realized edge-inclusion prior is the
edge prior reweighted by the per-graph normalizer, not the nominal edge prior
```

The choice between specifications is a modelling choice, not a tuning setting.
Use the fast route when runtime is the obstacle, and report which specification
produced the results. Use the default specification when the nominal edge
probability must equal the actual prior inclusion probability over graphs.

There is one practical consequence. Recovering the realized prior inclusion
probabilities, which `extract_prior_inclusion_probabilities()`,
`extract_inclusion_bf()` and `verdicts()` all require, depends on a table built
by simulation for the model size at hand. The table is computed once and cached
on disk, but it is not quick: at six variables it takes a few seconds, and at
fifty it takes longer than the fit itself. Posterior inclusion probabilities,
available from `summary()` or
`extract_posterior_inclusion_probabilities()`, require no such table.


# A mid-sized model, both ways

The first example has `r B$mid$p` variables and `r B$mid$n` observations,
drawn from a sparse Gaussian graphical model with `r B$mid$edges` edges among
`r B$mid$pairs` pairs. Both fits use the package defaults for run length
(`r formals(bgm)$warmup` warmup and `r formals(bgm)$iter` retained iterations)
and both run four chains on four threads. They differ in nothing else.

```{r, eval = FALSE}
default = bgm(y,
  variable_type = "continuous",
  chains = 4, cores = 4, seed = 2026
)

fast = bgm(y,
  variable_type = "continuous",
  precision_graph_prior = "joint",
  update_method = "gibbs",
  chains = 4, cores = 4, seed = 2026
)
```

```{r, echo = FALSE}
data.frame(
  route = c("hierarchical + nuts (default)", "joint + gibbs"),
  seconds = c(round(B$mid$default$elapsed, 1), round(B$mid$fast$elapsed, 1)),
  edges_found = c(B$mid$default$selected, B$mid$fast$selected),
  true_edges_found = c(B$mid$default$detected, B$mid$fast$detected),
  false_positives = c(
    B$mid$default$false_positive, B$mid$fast$false_positive
  ),
  max_rhat = round(c(B$mid$default$max_rhat, B$mid$fast$max_rhat), 3)
)
```

The default specification takes `r sec(B$mid$default$elapsed)` seconds and the
fast specification `r sec(B$mid$fast$elapsed)`, a factor of
`r sprintf("%.0f", B$mid$speedup)`. Both recovered `r B$mid$fast$detected` of
the `r B$mid$edges` true edges at the conventional 0.5 cut, and both are
converged.

The two edge sets agree on `r B$mid$agreement` of the `r B$mid$pairs` pairs.
That agreement is an observation about this data set, not a guarantee: the
sparser realized prior of the joint specification pulls verdicts toward
absence, and how far depends on how informative the data are. On data with
weaker signal, the two specifications will diverge sooner. The message of the
table is the seconds column: the recovery is the same, the runtime is not.


# Two hundred variables

The second example has `r B$high$p` variables and `r B$high$n` observations:
`r num(B$high$pairs)` candidate edges, `r B$high$edges` of them real. Only the
fast route was run.

```{r, eval = FALSE}
fit = bgm(y,
  variable_type = "continuous",
  precision_graph_prior = "joint",
  update_method = "gibbs",
  chains = 4, cores = 4, seed = 2026
)
```

```{r, echo = FALSE}
data.frame(
  variables = B$high$p,
  candidate_edges = B$high$pairs,
  seconds = round(B$high$elapsed, 1),
  edges_found = B$high$selected,
  true_edges_found = B$high$detected,
  false_positives = B$high$false_positive,
  max_rhat = round(B$high$max_rhat, 3)
)
```

The fit takes `r sec(B$high$elapsed)` seconds (about
`r sprintf("%.0f", B$high$elapsed / 60)` minutes) for four chains on four
threads at the package's default run length, estimating inclusion probabilities
for all `r num(B$high$pairs)` candidate edges. It found
`r B$high$detected` of the `r B$high$edges` true edges and
`r B$high$false_positive` pairs that are not edges, which on the
`r num(B$high$pairs - B$high$edges)` non-edges is a false-positive rate of
`r sprintf("%.1f%%", 100 * B$high$false_positive / (B$high$pairs - B$high$edges))`.

Nothing about reading the fit changes at this size:

```{r, eval = FALSE}
head(summary(fit)$indicator)
```

```{r, echo = FALSE}
round(B$high$head, 3)
```

The remaining question is how much posterior uncertainty remains across the
`r num(B$high$pairs)` candidate edges. The middle of the range contains
`r B$high$profile$pairs[3]` pairs
(`r sprintf("%.1f%%", 100 * B$high$profile$pairs[3] / B$high$pairs)` of the
candidates), so the pairs that need a closer look form a short list:

```{r, echo = FALSE}
knitr::kable(B$high$profile, col.names = c("inclusion probability", "pairs"))
```

The default route was not run at this size, so this vignette makes no claim
about what it would have cost. The measured comparison is the one at fifty
variables above.


# How it scales

Same generator, same `r B$high$n` observations, same four chains on four
threads, the fast route throughout.

```{r, echo = FALSE}
knitr::kable(
  data.frame(
    variables = B$scaling$p,
    candidate_edges = B$scaling$pairs,
    seconds = round(B$scaling$elapsed, 1)
  )
)
```

```{r, echo = FALSE, fig.alt = "Wall-clock seconds against the number of variables, on a logarithmic vertical axis, rising steeply from a fraction of a second at 25 variables to several minutes at 200."}
op = par(mar = c(4.2, 4.4, 1.2, 1.2), las = 1, bty = "n", cex = 0.9)
plot(B$scaling$p, B$scaling$elapsed,
  log = "y", type = "b", pch = 19, lwd = 2, cex = 1.1,
  col = "#2f6f9f",
  xlab = "Number of variables",
  ylab = "Seconds (four chains, four threads)",
  panel.first = grid(col = "grey92", lty = 1, lwd = 1)
)
text(B$scaling$p, B$scaling$elapsed,
  labels = sprintf("%.1f", B$scaling$elapsed),
  pos = c(4, 4, 4, 4, 2), offset = 0.6, cex = 0.8, col = "grey30"
)
par(op)
```

What drives that curve is the number of variables, not the number of
observations. Complete data enter the sampler only through the cross-product of
the data matrix, formed once before sampling starts; every draw after that
works on the precision matrix, whose size the variables alone set. More
observations change what the sampler is drawing from, not how much work a draw
costs. More variables change both.


# Where the fast route applies

`update_method = "gibbs"` applies only to the Gaussian graphical model.
Ordinal, Blume-Capel or mixed data are refused with an error:

```{r, error = TRUE}
bgm(Wenchuan[, 1:4], update_method = "gibbs")
```

Within continuous data it needs conjugate ingredients: a `normal_prior()` or
`cauchy_prior()` on the interactions, and `exponential_prior()` or
`gamma_prior()` on the precision diagonal. Those are the defaults for
`interaction_prior` and `precision_scale_prior`, so the usual call already
qualifies; `beta_prime_prior()` on the interactions does not. Edge selection is
supported for both slabs, and so is `edge_selection = FALSE`, which estimates
the precision matrix at a fixed graph.

`precision_graph_prior = "joint"` is separate from the sampler and applies to
continuous data under edge selection, in Gaussian graphical models and in the
continuous block of a mixed model. The two arguments compose but do not depend
on each other: the Gibbs sampler runs under the default hierarchical
specification, and the joint specification runs under `"nuts"`.

One combination is fast to sample but slow to set up. Under the joint
specification, a learned inclusion probability (`beta_bernoulli_prior()` or
`sbm_prior()`) needs the same simulated correction table described above,
because the constant that cancels for a fixed inclusion probability stops
cancelling once that probability is itself being updated. The table is built at
the first fit of a model configuration and cached, but at large numbers of
variables the wait for the table far exceeds the sampling time. For a fast run
at scale, keep `bernoulli_prior()`, the default.


# Background

The prior used here specifies the precision matrix entry by entry: an edge that
is absent has its entry fixed at exactly zero, and an edge that is present draws
its entry from a diffuse slab. The construction comes from the Bayesian
variable-selection literature [@george1993jasa_variable] and was brought to
Gaussian graphical models by @wang2015ba_scaling, whose column-wise block Gibbs
sampler is the ancestor of the one `update_method = "gibbs"` runs.

One difference from that sampler matters for runtime. Wang's spike is
continuous, a narrow normal around zero, so every column update redraws all of
the column's entries. `bgms` uses a point mass, so a column update works only
through the entries included in the current graph, and a sparser graph makes a
cheaper update.

The classical construction also avoids a graph-dependent normalizing constant.
The G-Wishart prior is the canonical example of the computational cost imposed
by such a constant
[@atay-kayis2005biometrika_monte; @mohammadi2023jasa_accelerating], but the cost
is not particular to the G-Wishart. `bgm()`'s hierarchical specification
retains a constant of the same kind so that the stated edge prior is exactly the
prior over graphs. The joint specification avoids that constant; the
hierarchical specification retains it to preserve the graph prior. This is the
choice exposed by `precision_graph_prior`.
@vogels2024jasa_bayesian survey the wider field and compare the available
samplers empirically.

`bgms` adds one thing to the classical construction. Specifying the precision
matrix entry by entry does not guarantee that the result is positive definite,
so the prior has to be truncated to the cone of positive-definite matrices, and
that truncation distorts it more and more as the number of variables grows
[@jewson2024biometrics_graphical]. `bgm()` counters this by tilting the prior by
a power `delta` of the determinant, which pushes mass away from the boundary of
the cone; `delta` defaults to half the logarithm of the number of variables, so
larger models are tilted harder. The tilt helps keep the prior usable at two
hundred variables rather than allowing the fit to collapse toward the empty
graph.

Scaling this prior is an active problem, and `bgms` is not alone in addressing
it. @sulem2025arxiv_bayesian give the closest treatment:
Metropolis-Hastings within block Gibbs on the same spike-and-slab Gaussian
graphical model, with mixing bounds that are dimension-free under suitable
conditions and a global proposal that adds or removes several edges in one
move. They report carrying exact Bayesian inference from roughly one hundred
variables to roughly one thousand. Their work is implemented in the CRAN
package **modelSelection** (version 1.0.7, 2026-05-16), which covers Bayesian
model selection and averaging for regression, generalized linear and additive
models, mixtures and graphical models. **ssgraph** (version 1.16, 2025-08-29)
is the other near neighbor, doing Bayesian structure learning for undirected
graphical models with spike-and-slab priors on continuous, discrete and mixed
data.

How does the fast route compare with these two packages? A measured comparison
is maintained on the package website's
[comparison page](https://bayesian-graphical-modelling-lab.github.io/bgms/guide/fast-ggm.html),
with the full protocol, seeded scripts, and complete tables
[in its repository](https://github.com/Bayesian-Graphical-Modelling-Lab/bgms-docs/tree/main/benchmarks/ggm-comparison).
It lives there rather than here because its results depend on the current
versions of the other packages, which change more quickly than `bgms` releases.

Because default run lengths differ by orders of magnitude, the comparison
measures the time required to reach a converged edge ranking, using the
convergence-cost metric of @vogels2024jasa_bayesian. On replicate datasets from
this vignette's generator, `bgms` reaches its converged ranking in seconds at
every tested sample size, a median 18 to 25 times sooner than ssgraph under a
matched edge prior. modelSelection converges within about a second, so its short
default run is converged rather than undersampled; it targets a sparser prior,
recovers somewhat fewer edges at a matched threshold, and cannot be seeded. The
three packages achieve similar converged ranking accuracy. They differ in how
quickly they reach it and in where their default prior distributions place the
selection threshold.

Readers whose problem is a single very large Gaussian graphical model and
nothing else should look at all three; what `bgms` adds around this route is
the rest of the package: ordinal, Blume-Capel and mixed models, group
comparison, prior sensitivity, and the determinant tilt that keeps the prior
usable as the number of variables grows.


# Reproducing the timings

Every timing in the worked `bgms` examples above was produced by
`data-raw/make-fast-ggm.R`, which is in the package repository and shows the
generator, seeds, and calls. Each of those fits ran on its own, with four chains
on four threads. The cross-package comparison uses a separate one-chain,
one-thread protocol whose scripts and complete results are linked in the
comparison section above.

```{r, echo = FALSE}
data.frame(
  field = c("CPU", "cores available", "threads used", "chains",
            "operating system", "R", "bgms", "run"),
  value = c(B$machine$cpu, B$machine$cores_total, B$machine$threads,
            B$machine$chains, B$machine$os, B$machine$r_version,
            B$machine$bgms_version, B$machine$date)
)
```

The timings are specific to this machine and these thread counts; the ratios
should transfer better than the absolute numbers.


# References
