---
title: "Missing data handling in bgms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing data}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---
<style>
body {
text-align: justify}
</style>

# Introduction

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("BUILD_VIGNETTE"), "true")
)
```

Starting with version 0.1.1 of bgms, the bgm function has a new feature called `na.action`. Currently there are two options for `na.action`. When `na.action = "listwise"`, the bgm function omits all rows containing NA's from the data in the analysis (i.e., these rows are deleted from the data). This was the default procedure used by bgms prior to version 0.1.1. 

Conversely, if `na.action = "impute"`, the NAs are imputed during the underlying MCMC procedure. That is, in each iteration of the MCMC algorithm, missing values are imputed with random samples from the model based on the current state of the parameter values and the remaining observed data. After the missing data are imputed, we have a complete data set and the algorithm proceeds as usual. 

In this vignette, we examine the impact of missing data for simulated and real data on PTSD symptoms of 362 Chinese adults who survived the Wenchuan earthquake but tragically lost a child [@McNallyEtAl_2015]. The data come from a 17-question survey in which participants rated how much each symptom bothered them in the past month on a scale from "not at all" to "extremely. First, we examine the data.

# Example 1 -- Wenchuan data
```{r setup}
library(bgms)

sum(is.na(Wenchuan))
x = na.omit(Wenchuan)
nrow(x)
```

Here we see that there are 22 missing observations in 18 of the 362 rows. In total, less than one percent of the data is missing. However, when we use listwise deletion, we omit almost 5% of the data. 

## Analysis
Here we analyze how much the results differ between the two `na.action` options. To do this, we fit the model with `na.action = "listwise"` and with `na.action = "impute"`.

```{r message = FALSE, warning = FALSE, cache = TRUE}
fit.listwise = bgm(x = Wenchuan, na.action = "listwise")
fit.impute = bgm(x = Wenchuan, na.action = "impute")
```
Let's look at how the estimated thresholds, interaction parameters, and inclusion probabilities compare between the `na.action` options. To do this, we make a small panel plot comparing the posterior means in each setting. First, we compute the means.

```{r message = FALSE, warning = FALSE, fig.width = 4, fig.height = 12}
par(mfrow = c(3, 1), mar = c(6, 5, 1, 1))
plot(fit.listwise$thresholds, fit.impute$thresholds, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("Thresholds | Listwise", side = 1, line = 3, cex = .9)
mtext("Thresholds | Impute", side = 2, line = 3, cex = .9)

plot(fit.listwise$interactions, fit.impute$interactions, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("Interactions | Listwise", side = 1, line = 3)
mtext("Interactions | Impute", side = 2, line = 3)


plot(fit.listwise$gamma, fit.impute$gamma, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("PIP | Listwise", side = 1, line = 3)
mtext("PIP | Impute", side = 2, line = 3)
```

In the plot above, the two `na.action` options do not seem to affect the estimation of the threshold parameters too much for this example. There are some differences, but they are relatively small. We see some larger differences between the estimated interactions. For example, some interactions were estimated to be nonzero with `na.action = "listwise"` that were estimated to be zero with `na.action = "impute"`. The largest differences are observed for the posterior inclusion probabilities, whose estimates vary greatly between the two options.

# Example 2 -- Simulated data with 22 missings

From the analysis above, it is difficult to determine whether the different estimates, especially of the inclusion probabilities, are the result of the 4% missing data in the `na.action = "listwise"` case, but not in the `na.action = "impute"` case. To verify that the imputation methods work correctly, we run a small simulation.

First, we generate new data according to the model estimated from the Wenchuan data using `na.action = "listwise"`.

```{r message = FALSE, warning = FALSE}
interactions = fit.listwise$interactions
thresholds = fit.listwise$thresholds
x = mrfSampler(no_states = nrow(Wenchuan),
               no_nodes = ncol(Wenchuan), 
               no_categories = 4, 
               interactions = interactions, 
               thresholds = thresholds,
               iter = 1e3)
```

Now we generate 22 missing data points, just like in the original Wenchuan data.

```{r message = FALSE, warning = FALSE}
y = x
for(mis in 1:22) {
  stop = FALSE
  while(stop == FALSE) {
    i = sample(1:nrow(x), 1)
    j = sample(1:ncol(x), 1)

    if(!is.na(y[i, j])) {
      stop = TRUE
      y[i,j] = NA
    }
  }
}
```

We can now estimate the model for the complete data case, and we can use listwise deletion and imputation for the missing data case.

```{r message = FALSE, warning = FALSE}
newfit.obs = bgm(x = x)
newfit.listwise = bgm(x = y, na.action = "listwise")
newfit.impute = bgm(x = y, na.action = "impute")
```
Let's first look at the estimated thresholds and make a small panel plot comparing the posterior medians in each setting.

```{r message = FALSE, warning = FALSE, fig.width = 5, fig.height = 7}
par(mfrow = c(3, 1), mar = c(6, 5, 1, 1))
plot(newfit.obs$thresholds, newfit.listwise$thresholds, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("Thresholds | Complete", side = 1, line = 3, cex = .9)
mtext("Thresholds | Listwise", side = 2, line = 3, cex = .9)

plot(newfit.obs$thresholds, newfit.impute$thresholds, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("Thresholds | Complete", side = 1, line = 3, cex = .9)
mtext("Thresholds | Impute", side = 2, line = 3, cex = .9)

plot(newfit.listwise$thresholds, newfit.impute$thresholds, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("Thresholds | Listwise", side = 1, line = 3, cex = .9)
mtext("Thresholds | Impute", side = 2, line = 3, cex = .9)
```

In the plot above, similar to the previous example using the Wenchuan data, we see that the two `na.action` options do not appear to affect the estimation of the threshold parameters. Next, we look at the estimated interaction effects.  

```{r message = FALSE, warning = FALSE, fig.width = 5, fig.height = 7}
par(mfrow = c(3, 1), mar = c(6, 5, 1, 1))
plot(newfit.obs$interactions, newfit.listwise$interactions, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("Interactions | Complete", side = 1, line = 3, cex = .9)
mtext("Interactions | Listwise", side = 2, line = 3, cex = .9)

plot(newfit.obs$interactions, newfit.impute$interactions, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("Interactions | Complete", side = 1, line = 3, cex = .9)
mtext("Interactions | Impute", side = 2, line = 3, cex = .9)

plot(newfit.listwise$interactions, newfit.impute$interactions, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("Interactions | Listwise", side = 1, line = 3, cex = .9)
mtext("Interactions | Impute", side = 2, line = 3, cex = .9)
```

In the plot above, we see that the two `na.action` options do not seem to affect the estimation of the interaction parameters, contrary to what we saw for the previous Wenchuan data example. We expect the same for the inclusion probabilities.

```{r message = FALSE, warning = FALSE, fig.width = 5, fig.height = 7}
par(mfrow = c(3, 1), mar = c(6, 5, 1, 1))
plot(newfit.obs$gamma, newfit.listwise$gamma, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("PIP | Complete", side = 1, line = 3, cex = .9)
mtext("PIP | Listwise", side = 2, line = 3, cex = .9)

plot(newfit.obs$gamma, newfit.impute$gamma, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("PIP | Complete", side = 1, line = 3, cex = .9)
mtext("PIP | Impute", side = 2, line = 3, cex = .9)

plot(newfit.listwise$gamma, newfit.impute$gamma, xlab = "", ylab = "", axes = FALSE)
abline(0,1)
axis(1)
axis(2, las = 1)
mtext("PIP | Listwise", side = 1, line = 3, cex = .9)
mtext("PIP | Impute", side = 2, line = 3, cex = .9)
```

There appears to be a bit more variability in the estimated posterior inclusion probabilities for the `na.action = "listwise"` case than for the `na.action = "impute"` case. Although there is still large variability for the imputation case, this variability appears to be mostly due to the sampling variability of the underlying MCMC procedure and largely disappears as the number of iterations increases.

## References