---
title: "Checking Prior Sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking Prior Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

# Why the scale is the reader's problem

An analysis with `bgm()` states its assumptions as prior distributions, so
the natural question is whether the conclusions depend on them. The
interaction (pairwise) prior distribution is the one to worry about first,
because its scale encodes how large you expect edge weights to be, and that
expectation feeds directly into every inclusion verdict.

The scale is interpretable. A pairwise interaction in `bgm()` is half the log
odds ratio between adjacent response categories, so the default
`normal_prior(scale = 1)` places about two thirds of its prior mass on odds
ratios between roughly 1/7 and 7. A reader who thinks edges are typically
weaker wants a narrower scale; a reader who expects strong dependencies wants
a wider one. If the inclusion verdicts survive that disagreement, they belong
to the data. If they move, they belong to the scale, and the analysis should
say so.

```{r, eval = FALSE}
library(bgms)
data = Wenchuan[, 1:9]
```

```{r, include = FALSE}
library(bgms)
data = Wenchuan[, 1:9]
```

# The brute-force check

The direct check reruns the analysis under a few reasonable alternatives and
compares the verdicts.

```{r, eval = FALSE}
fit_default = bgm(data, seed = 1234)
fit_narrow  = bgm(data, interaction_prior = normal_prior(scale = 0.5), seed = 1234)
fit_wide    = bgm(data, interaction_prior = cauchy_prior(scale = 2.5), seed = 1234)
```

What matters is not that every number is identical; it is whether edges change
categories. A finding that survives is one whose verdict is stable across the
reasonable prior distributions.

One caution carries over from convergence checking. A verdict whose inclusion
Bayes factor sits within Monte Carlo error of an evidence threshold can change
category between prior settings, or even between reruns of the same setting,
without any real sensitivity being present. A move that stays inside Monte
Carlo error of a threshold says more about the threshold than about the prior
distribution.

The brute-force check is honest but coarse: it visits a handful of scales, and
it cannot tell a genuine move from Monte Carlo wobble without extra reruns.

# The principled check

A more systematic check treats the scale itself as the quantity in question
and asks, for every edge, over which range of scales its verdict holds. Supply
an `interaction_scale_prior` to make the slab scale a sampled parameter,
centered on the chosen scale. The scale is written `s = s0 * u`, where `s0` is
the scale from `interaction_prior` and `u` follows a mean-one prior
distribution. The default recommendation is `gamma_prior(shape = 2, rate = 2)`,
which places 95% of its mass on roughly 0.12 to 2.8 times the chosen scale.

```{r, eval = FALSE}
fit = bgm(data, interaction_scale_prior = gamma_prior(2, 2), seed = 1234)
```

```{r, include = FALSE}
fit = bgm(data, interaction_scale_prior = gamma_prior(2, 2),
          iter = 2000, warmup = 1000, chains = 2,
          display_progress = "none", verbose = FALSE)
```

This single fit answers the sensitivity question two ways. First, its raw
inclusion probabilities already average the fixed-scale answer over the
posterior uncertainty about the scale, so the fit reports a scale-averaged
verdict with no post-processing at all. Second, each edge's whole inclusion
Bayes factor curve, as a function of the scale, is recovered from the same fit
by reweighting.

`prior_sensitivity_check()` reports both.

```{r}
ps = prior_sensitivity_check(fit)
ps
```

The `mover` edges are the ones whose verdict changes with the scale, or whose
chosen-scale verdict differs from the scale-averaged one. The `plot` method
draws the inclusion Bayes factor curves against the relative scale, with the
undecided band shaded and the chosen scale marked. Mover edges are drawn solid
and colored by their chosen-scale verdict; stable edges are faint. Thin
per-chain curves are drawn underneath: where chains disagree about a crossing,
the move is Monte Carlo wobble, not prior sensitivity. The lower panel is a
forest of the scale-stability intervals for the mover edges.

```{r}
plot(ps)
```

The per-edge table carries the marginalized inclusion probability and Bayes
factor, the marginalized and chosen-scale verdicts, and each edge's stability
interval in relative units.

```{r}
head(ps$edges)
```

# What to report

Report the settings and the outcome in the same register as the rest of the
analysis:

- the hyperprior used and the evidence thresholds,
- whether any edge moves between presence and absence anywhere on the grid (a
  presence-to-absence swap is the one result a sensitivity analysis must not
  hide),
- that the movers are confined to the undecided boundary, if they are,
- the stability intervals for the edges the analysis discusses.

An edge that is included or excluded in every draw is scale-robust by
construction, and the check reports it as showing presence, or absence, at
every scale considered rather than as a number.

# How the single fit does it

The construction rests on one structural fact: the scale enters the model only
through the slab, never through the likelihood and never through the edge
prior distribution [@bartos2026]. Two consequences follow.

The scale-averaged verdict is free. By the law of total probability the raw
inclusion probability of the extended fit is the fixed-scale inclusion
probability averaged over the scale posterior. Its inclusion Bayes factor
divides the posterior inclusion odds by the prior inclusion odds, taken from
`extract_prior_inclusion_probabilities()` so that a non-even edge prior is
handled correctly.

The fixed-scale curve is a reweighting. Conditioning the joint posterior on a
scale value is the same as reweighting the sampled draws, and the weight of a
draw is the conditional density of the scale given that draw's included
interactions,

$$
w_t(s) \;\propto\; \pi(s) \prod_{e:\, \gamma_e^{(t)} = 1}
  \phi\!\left(\theta_e^{(t)};\, 0,\, s^2\right),
$$

normalized over the scale by one-dimensional quadrature. The likelihood and
the edge prior cancel because neither involves the scale, so no likelihood
evaluation is needed after the fit, and the same weight matrix prices every
edge's curve at once. The conditional inclusion probability at scale `s` is
then the weighted share of draws that include the edge.

The estimator is a self-normalized importance-sampling average, so it is
trustworthy only where enough draws carry weight. The reliability at a scale
is measured by the effective sample size of the weights there, and the curve
is returned as `NA` where that effective sample size falls below `ess_floor`
(400 by default) rather than extrapolated. Because the scale is informed only
by the included interactions, in a sparse network its posterior stays close to
the hyperprior, so centering the hyperprior on the chosen scale puts the
reliability exactly where the sensitivity question is asked.

The accuracy of the reweighting has been checked against exact enumeration in
a conjugate testbed: over scales between one third and three times the chosen
scale, the largest error in $\log_{10}$ inclusion Bayes factor was about 0.04
in the evidence-of-absence region, 0.05 near even odds, and 0.14 near the
presence threshold, small against the two $\log_{10}$ units that separate the
evidence thresholds.

# References
