---
title: "Checking Prior Sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking Prior Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

# Why the scale is the reader's problem

An analysis with `bgm()` states its assumptions as prior distributions, so
the natural question is whether the conclusions depend on them. The
interaction (pairwise) prior distribution is the one to worry about first,
because its scale encodes how large you expect edge weights to be, and that
expectation feeds directly into every inclusion verdict.

The scale is interpretable. A pairwise interaction in `bgm()` is half the log
odds ratio between adjacent response categories, so the default
`normal_prior(scale = 1)` places about two thirds of its prior mass on odds
ratios between roughly 1/7 and 7. A reader who thinks edges are typically
weaker wants a narrower scale; a reader who expects strong dependencies wants
a wider one. If the inclusion verdicts survive that disagreement, they belong
to the data. If they move, they belong to the scale, and the analysis should
say so.

```{r, eval = FALSE}
library(bgms)
data = Wenchuan
```

```{r, include = FALSE}
library(bgms)
data = Wenchuan
```

# The check

`prior_sensitivity_check()` is that comparison, automated. It takes a fitted
model, refits it at a small grid of slab scales --- narrow, chosen, and wide
--- and classifies every edge at every scale. No special fit is required: it
works on any `bgm()` fit made with edge selection.

```{r, eval = FALSE}
fit = bgm(data)
ps = prior_sensitivity_check(fit)
ps
```

```
#> PLACEHOLDER-PRINT-OUTPUT
```

The scales are multipliers of the fit's chosen scale, so one default
(`c(0.5, 1, 2.5)`) serves `normal_prior(1)` and `cauchy_prior(2.5)` alike. The
category counts are the headline: most verdicts are robust, and the edges
whose verdict genuinely depends on the scale are named, with their verdict at
each scale. The per-scale verdict table below them shows the expected drift:
as the slab widens, evidence for inclusion weakens and edges move from
undecided toward absence, since a wider slab penalizes the extra parameter an
edge costs.

## Run-to-run noise

Verdict-identity comparisons are flaky by construction: an inclusion Bayes
factor that sits within Monte Carlo error of a threshold can change category
between reruns of the *same* prior, with no real sensitivity present. The
multiplier `1` is included for exactly this reason. Its refit is a repeat run
of the chosen-scale analysis, and the spread of $|\Delta \log_{10} \mathrm{BF}|$
between the chosen-scale refit and this repeat is the empirical noise band
(its 95th percentile is the "run-to-run noise band" in the report's Details
line, computed over edges whose evidence is near the decision thresholds).

An edge is reported as "changed, beyond run-to-run noise" only if its verdict
changes across the grid *and* its $\log_{10}$ Bayes-factor swing exceeds that
band (and its own Monte Carlo error, and a small tolerance). An edge that
flips inside the band is "changed, within run-to-run noise": the check
refuses to call it a move. This is what keeps a robustness tool from crying
wolf on boundary edges. In the `$edges$mover` column these categories are
stored as `stable`, `indistinguishable-from-wobble`, and
`moved-beyond-wobble`.

```{r, eval = FALSE}
plot(ps)
```

The plot's title states the answer. Each edge's evidence line runs across the
grid, the undecided band is shaded and the verdict zones are labeled; the
edges whose verdict depends on the scale are colored and labeled by name, and
everything else is muted background.

The per-edge table carries the chosen-scale verdict and $\log_{10}$ Bayes
factor with its Monte Carlo standard error, the per-scale verdict columns, the
stability range, the mover category, and an `insufficient` flag for edges the
refit cannot certify (its chains disagree, or the Bayes-factor band straddles a
threshold); the report prints these as "not certifiable".

```{r, eval = FALSE}
head(ps$edges)
```

## The chosen scale versus the data

One line of the report costs no refit: how the chosen scale compares with the
size of the estimated interactions (the root-mean-square of the included
interactions in the original fit, with an approximate standard error). When
the chosen scale is far wider or narrower than that --- more than about a
factor of two --- the report says so, because inclusion verdicts near the
boundary then depend on the choice, and the grid shows by how much.

## Cost and samplers

With warm starts, each refit begins from the original fit's per-chain end
state; an ordinal fit sampled with NUTS also carries the adapted step size and
mass matrix, so a short warmup suffices. The scales are refit one after another,
each running its chains in parallel across `cores`, and on Wenchuan the whole
three-scale check (plus the chosen-scale repeat) costs about one original
fit of wall-clock. A fit sampled with adaptive Metropolis or Gibbs refits with
the same sampler and costs about one fit per scale; when that makes the check
cost more than about a minute, it suggests `refit_sampler = "nuts"`, which is
sound because every comparison runs refit-against-refit --- the original fit
never enters a comparison. Continuous (GGM) and mixed-MRF fits refit cold.

Because warm starts place every chain near the chosen-scale posterior,
between-chain dispersion is reduced by construction, so split-$\hat R$ is a
weakened diagnostic at the refits; the per-refit convergence gate therefore also
leans on the effective sample size and on per-chain verdict agreement. A refit
that fails its gate is reported as unusable rather than folded into the verdicts.
The check thus doubles as a convergence audit of the original analysis: an
undermixed original run shows up as a refit that cannot reproduce its own
chosen-scale verdicts within Monte Carlo error.

# References
