---
title: "Checking Prior Sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking Prior Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
csl: apa.csl
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

# Why the scale is the reader's problem

An analysis with `bgm()` states its assumptions as prior distributions, so
the natural question is whether the conclusions depend on them. The
interaction (pairwise) prior distribution is the one to worry about first,
because its scale encodes how large you expect edge weights to be, and that
expectation feeds directly into every inclusion verdict.

The scale is interpretable. A pairwise interaction in `bgm()` is half the log
odds ratio between adjacent response categories, so the default
`normal_prior(scale = 1)` places about two thirds of its prior mass on odds
ratios between roughly 1/7 and 7. A reader who thinks edges are typically
weaker wants a narrower scale; a reader who expects strong dependencies wants
a wider one. If the inclusion verdicts survive that disagreement, they belong
to the data. If they move, they belong to the scale, and the analysis should
say so.

```{r, eval = FALSE}
library(bgms)
data = Wenchuan
```

```{r, include = FALSE}
library(bgms)
data = Wenchuan
```

# The check

`prior_sensitivity_check()` is that comparison, automated. It takes a fitted
model, refits it at a small grid of slab scales --- narrow, chosen, and wide
--- and classifies every edge at every scale. No hyperprior and no special fit
are required: it works on any `bgm()` fit made with edge selection.

```{r, eval = FALSE}
fit = bgm(data)
ps = prior_sensitivity_check(fit)
ps
```

```
#> Prior sensitivity of edge inclusion verdicts
#>
#> Chosen scale 1; the data prefer approximately 0.15 [0.12, 0.18].
#> Evidence thresholds: presence >= 10, absence <= 0.1.
#>
#> Verdict counts by slab scale (multiplier x chosen scale):
#>           0.5x  1x 2.5x
#> presence    37  35   31
#> undecided   60  44   31
#> absence     39  57   74
#>
#> Scale sensitivity (wobble q95 = 0.28 log10 BF from the s0 replicate):
#>   stable:                        118
#>   indistinguishable-from-wobble:  13
#>   moved-beyond-wobble:             5
#>
#> Refit sampler: nuts (warm-started). Whole check: 205 s (4 refits).
```

The scales are multipliers of the fit's chosen scale, so one default
(`c(0.5, 1, 2.5)`) serves `normal_prior(1)` and `cauchy_prior(2.5)` alike. The
verdict-count table is the headline: as the slab widens, evidence for inclusion
weakens and edges drift from undecided toward absence --- the expected direction,
since a wider slab penalizes the extra parameter an edge costs.

## The wobble yardstick

Verdict-identity comparisons are flaky by construction: an inclusion Bayes
factor that sits within Monte Carlo error of a threshold can change category
between reruns of the *same* prior, with no real sensitivity present. The
multiplier `1` is included for exactly this reason. Its refit is a replicate of
the chosen-scale analysis, and the spread of $|\Delta \log_{10} \mathrm{BF}|$
between the chosen-scale refit and this replicate is the empirical wobble
yardstick (its 95th percentile is printed above).

An edge is called `moved-beyond-wobble` only if its verdict changes across the
grid *and* its $\log_{10}$ Bayes-factor swing exceeds that yardstick (and its
own Monte Carlo error, and a small tolerance). An edge that flips inside the
wobble is `indistinguishable-from-wobble`: the check refuses to call it a move.
This is what keeps a robustness tool from crying wolf on boundary edges.

```{r, eval = FALSE}
plot(ps)
```

The plot draws one $\log_{10}$ Bayes-factor line per edge across the grid, the
undecided band shaded and the chosen scale marked; `moved-beyond-wobble` edges
are solid and colored by verdict, everything else faint. The lower panel is a
forest of the scale ranges over which each mover's verdict holds.

The per-edge table carries the chosen-scale verdict and $\log_{10}$ Bayes
factor with its Monte Carlo standard error, the per-scale verdict columns, the
stability range, the mover category, and an `insufficient` flag for edges the
refit cannot certify (its chains disagree, or the Bayes-factor band straddles a
threshold).

```{r, eval = FALSE}
head(ps$edges)
```

## What the report opens with

The first line is the single most useful sentence the check produces, and it
costs no refit: the scale the data prefer. It is the root-mean-square of the
included interactions in the original fit, with an approximate standard error.
When the data prefer a scale far from the one you chose --- here about `0.15`
against a chosen `1` --- the inclusion verdicts near the boundary will be scale
sensitive, and the grid shows by how much.

## Cost and samplers

With warm starts, each refit begins from the original fit's per-chain end
state; an ordinal fit sampled with NUTS also carries the adapted step size and
mass matrix, so a short warmup suffices. The scales are refit one after another,
each running its chains in parallel across `cores`, and on Wenchuan the whole
three-scale check (plus the chosen-scale replicate) costs about one original
fit of wall-clock. A fit sampled with adaptive Metropolis or Gibbs refits with
the same sampler and costs about one fit per scale; the check then suggests
`refit_sampler = "nuts"` for speed, which is licensed because every comparison
runs refit-against-refit. Continuous (GGM) and mixed-MRF fits refit cold.

Because warm starts place every chain near the chosen-scale posterior,
between-chain dispersion is reduced by construction, so split-$\hat R$ is a
weakened diagnostic at the refits; the per-refit convergence gate therefore also
leans on the effective sample size and on per-chain verdict agreement. A refit
that fails its gate is reported as unusable rather than folded into the verdicts.
The check thus doubles as a convergence audit of the original analysis: an
undermixed original run shows up as a refit that cannot reproduce its own
chosen-scale verdicts within Monte Carlo error.

# Letting the data choose the scale

Robustness ("do my verdicts survive a reasonable disagreement about the scale?")
is a different question from scale learning ("what scale do the data imply?").
For the second, supply an `interaction_scale_prior` to make the slab scale a
sampled parameter centered on the chosen scale, written $s = s_0 u$ with $u$
following a mean-one prior. The recommended default is
`gamma_prior(shape = 2, rate = 2)`.

```{r, eval = FALSE}
fit_learned = bgm(data, interaction_scale_prior = gamma_prior(2, 2))
ps_learned = prior_sensitivity_check(fit_learned)
```

When the fit carries a scale hyperprior, the report adds a clearly labeled
learned-scale block: the realized-versus-nominal scale interval, and the
scale-averaged verdict counts. On Wenchuan the data pull the scale down to about
`0.15`, and at that learned scale the verdicts read roughly `40 / 96 / 0`
(presence / undecided / absence) --- almost nothing is called absent --- against
`35 / 44 / 57` at the chosen scale of `1`. That contrast is the cautionary
point: the learned scale answers "what scale fits?", not "is my verdict robust?".
The robustness grid above is the tool for the second question, and the two
should not be conflated.

# References
